---
title: "Processing IDOT Crash Data (2022–2024)"
subtitle: "From yearly CSV exports to a clean sf data frame"
date: today
format:
  html:
    toc: true
    code-fold: false
    code-tools: true
    df-print: paged
execute:
  warning: true
  message: true
params:
  data_dir: "path/to/crash_csvs"
  years: [2022, 2023, 2024]
---

## Overview

This document reads Illinois Department of Transportation (IDOT) crash
records exported as CSV files from the IDOT open data portal. It stacks
several years, converts all column names to lowercase, cleans key fields,
and converts the result to an `sf` point data frame. It also produces a
subset of pedestrian and bicycle crashes.

Each yearly CSV should sit in the folder set by `params$data_dir`, with
the year somewhere in its file name (for example, `Crashes_2023.csv`).

## Setup

```{r}
#| label: setup
#| message: false

library(readr)
library(dplyr)
library(sf)
library(lubridate)
library(purrr)

data_dir <- params$data_dir
years    <- unlist(params$years)
```

## Functions

### Read one year

`read_crash_year()` finds the CSV for a given year and reads every column
as character. Reading as character lets years with different column types
be stacked without errors; types are restored later with `type_convert()`.
Column names are converted to lowercase here.

```{r}
#| label: fn-read

read_crash_year <- function(year, dir) {
  file <- list.files(dir, pattern = paste0(year, ".*\\.csv$"),
                     full.names = TRUE, ignore.case = TRUE)

  if (length(file) == 0) {
    warning("No CSV found for ", year, " in ", dir, "; skipping.")
    return(NULL)
  }
  if (length(file) > 1) {
    warning("Multiple CSVs match ", year, "; using ", basename(file[1]))
    file <- file[1]
  }

  read_csv(file, col_types = cols(.default = col_character()),
           show_col_types = FALSE) %>%
    rename_with(tolower) %>%
    mutate(source_file = basename(file))
}
```

### Process multiple years

`process_crashes()` stacks the years, cleans the data, and returns an `sf`
object. Its arguments are:

- `years`: a vector of years, such as `2022:2024`.
- `dir`: the folder containing the yearly CSVs.
- `crs_out`: the output coordinate reference system. Use 4326 for
  longitude/latitude, or 3435 for Illinois State Plane East in feet
  (3436 for Illinois West).
- `ped_bike_only`: if `TRUE`, returns only pedestrian and bicycle crashes.

A few cleaning choices are worth noting:

- **Coordinates.** For each row, the ArcGIS `x`/`y` columns are used when
  they fall inside Illinois. Otherwise the function falls back to IDOT's
  `tscrashlongitude`/`tscrashlatitude`. Records with neither are dropped
  and counted by year.
- **Dates.** The crash date is built from `crashyr`, `crashmonth`, and
  `crashday` rather than parsed from `crashdate`, because exported dates
  are often stored in UTC and can shift late-evening crashes to the next
  day.
- **Pedestrian and bicycle crashes.** These are identified from
  `typeoffirstcrash`. This field records only the *first* harmful event,
  so crashes where a pedestrian or cyclist was struck later in the crash
  sequence are not captured.
- **KSI.** A crash is flagged as killed or seriously injured when it has
  at least one fatality or A-level (incapacitating) injury.

```{r}
#| label: fn-process

process_crashes <- function(years, dir, crs_out = 4326,
                            ped_bike_only = FALSE) {

  crash_raw <- years %>%
    map(read_crash_year, dir = dir) %>%
    compact() %>%
    bind_rows() %>%
    type_convert(col_types = cols())

  in_il <- function(lon, lat) {
    !is.na(lon) & !is.na(lat) &
      between(lon, -91.6, -87.0) & between(lat, 36.9, 42.6)
  }

  crash_df <- crash_raw %>%
    mutate(
      lon = if_else(in_il(x, y), x, tscrashlongitude),
      lat = if_else(in_il(x, y), y, tscrashlatitude),

      crash_date = make_date(crashyr, crashmonth, crashday),
      crash_dt   = crash_date + hours(crashhour),

      ped_bike = case_when(
        typeoffirstcrash == "Pedestrian"   ~ "Pedestrian",
        typeoffirstcrash == "Pedalcyclist" ~ "Bicycle",
        TRUE                               ~ NA_character_
      ),
      ksi = totalfatals > 0 | ainjuries > 0
    ) %>%
    select(-any_of("crash_hour"))

  crash <- crash_df %>%
    filter(in_il(lon, lat)) %>%
    st_as_sf(coords = c("lon", "lat"), crs = 4326, remove = FALSE)

  dropped <- crash_df %>%
    filter(!in_il(lon, lat)) %>%
    count(crashyr, name = "dropped_no_coords")
  if (nrow(dropped) > 0) {
    message("Records dropped for missing/invalid coordinates:")
    print(dropped)
  }

  if (ped_bike_only) crash <- filter(crash, !is.na(ped_bike))

  st_transform(crash, crs_out)
}
```

## Process the data

### All crashes

```{r}
#| label: run-all

crash <- process_crashes(years, dir = data_dir)

crash %>%
  st_drop_geometry() %>%
  count(crashyr, name = "crashes")
```

### Pedestrian and bicycle crashes

This subset is projected to Illinois State Plane East so that distances
and buffers are in feet.

```{r}
#| label: run-pedbike

ped_bike <- process_crashes(years, dir = data_dir,
                            crs_out = 3435, ped_bike_only = TRUE)

ped_bike %>%
  st_drop_geometry() %>%
  count(crashyr, ped_bike, crashseverity)
```

## Checks

Confirm that the severity label and code columns line up, and that
`crashday` is the day of the month (values 1–31) rather than a day of
the week.

```{r}
#| label: checks

crash %>% st_drop_geometry() %>% count(crashseveritycd, crashseverity)

crash %>% st_drop_geometry() %>% summarise(min_day = min(crashday),
                                           max_day = max(crashday))
```

A quick map of pedestrian and bicycle crashes:

```{r}
#| label: map
#| fig-width: 7
#| fig-height: 8

plot(st_geometry(ped_bike), pch = 20, cex = 0.3,
     col = ifelse(ped_bike$ped_bike == "Pedestrian", "firebrick", "steelblue"),
     main = "Pedestrian (red) and bicycle (blue) crashes")
```

## Save

```{r}
#| label: save
#| eval: false

st_write(crash, "idot_crashes_2022_2024.gpkg", delete_dsn = TRUE)
saveRDS(ped_bike, "ped_bike_2022_2024.rds")
```

## Notes

- Police-reported crash data undercounts pedestrian and bicycle crashes,
  especially cyclist falls and minor injuries with no police report.
- Injury counts (`totalfatals`, `ainjuries`, and so on) cover everyone
  involved in the crash, not only the pedestrian or cyclist.
- Variable definitions are in IDOT's *Dictionary of Data Elements:
  Illinois Traffic Crash Report (SR 1050) and Crash Information System
  (CIS)*.
