---
title: "Data Preparation for Phase 1 — Dataset Development"
subtitle: "Street Design as a Public Health Determinant: Crosswalk Geometry and Pedestrian Injury Risk Across Cook County"
date: today
format:
  html:
    toc: true
    toc-depth: 2
    code-fold: true
    number-sections: false
    theme: cosmo
    embed-resources: true
    warning: false
editor: source
execute:
  eval: false
editor_options:
  chunk_output_type: console
---

## Purpose

This document builds the `sf` data tables that support [`phase01.qmd`](phase01.qmd)'s dataset-development tasks: the countywide study-area geography, street intersections, the street/speed-limit network, pedestrian fatality records (NHTSA FARS), and crash records (IDOT). It ends by saving the resulting tables to `data/phase1_dataprep.RData`, which `phase01.qmd` loads (a subset of) to build an interactive map for data exploration.

## Activate R Libraries

```{r}
#| label: r-packages
#| eval: true

library(sf)
library(reticulate)
library(dplyr)
library(tidyr)
library(purrr)
library(readr)
library(glue)
library(cli)
library(tigris)
library(units)
library(osmextract)
library(lubridate)
library(stringr)
library(tidycensus)
library(sociome)
library(tidyr)
library(arcgislayers)
```

## Study Area Geography

Combines the City of Chicago's 77 community areas with suburban Cook County's incorporated municipalities (≥4 sq mi) into a single, internally consistent geography (`study_area`, N ≈ 200), identified by `geoid` and `name`, with a `type` field distinguishing the two as either `community area` or `municipality`.

### Cook County boundary
```{r}
#| label: study-area

# download US Census TIGER/Line Files using tigris package
counties_il <- counties(state = "17", cb = TRUE)

cc_boundary <- counties_il %>%
  filter(GEOID == "17031") %>%
  st_transform(3435) %>%
  select(geoid_county = GEOID, name_county = NAME,
         area_land = ALAND, area_water = AWATER) %>%
  mutate(
    area_land_sqmi  = set_units(set_units(area_land, m^2), mi^2),
    area_water_sqmi = set_units(set_units(area_water, m^2), mi^2),
    area_sqmi = area_land_sqmi + area_water_sqmi
  ) %>%
  select(-area_land, -area_water)

st_write(cc_boundary, "data/boundaries/cc_boundary.gpkg", delete_dsn = TRUE)
```

### Councils of mayors boundaries
```{r}
# import, process councils of Mayors boundaries for Cook County
cogs_boundaries <- st_read(dsn="data/boundaries/councils_of_mayors.shp") %>% 
  rename_with(tolower) %>% 
  select(council,
         name = fullname) %>% 
  st_transform(crs = 3435) 

st_write(cogs_boundaries, "data/boundaries/cogs_boundaries.gpkg", delete_dsn = TRUE)
```

### Study areas boundaries (Chicago community area and Suburban municipalities)
```{r}
# download community areas from Chicago Health Atlas
community_areas <- st_read(
  "https://metopio.blob.core.windows.net/lalage/atlas/3/shapes/communityareas.topo.json",
  quiet = TRUE
) %>%
  st_as_sf() %>%
  st_set_crs(4326) %>%
  st_transform(3435) %>%
  rename(geoid = id) %>%
  mutate(
    geoid = str_replace(geoid, "1714000-", ""),
    location = "Chicago, IL",
    type = "community area"
  )

# download municipalities from Cook County Health Atlas
places_cook_county <- st_read(
  "https://metopio.blob.core.windows.net/lalage/atlas/6/shapes/SCC-place.topo.json",
  quiet = TRUE
) %>%
  st_as_sf() %>%
  st_set_crs(4326) %>%
  st_transform(3435) %>%
  rename(geoid = id) %>%
  mutate(
    sqmi = as.numeric(st_area(geometry)) * 0.0000003861,
    location = "suburban Cook County",
    type = "municipality"
  ) %>%
  filter(sqmi >= 4) %>%
  select(-sqmi)

study_area <- places_cook_county %>%
  bind_rows(community_areas) %>%
  st_as_sf() %>%
  st_make_valid()

st_write(study_area, "data/boundaries/study_area.gpkg", delete_dsn = TRUE)

```

## Street Intersections

Street intersections are derived offline from an Illinois-wide OpenStreetMap extract, downloaded from [GeoFabrik](https://download.geofabrik.de). The extract is first clipped to the Cook County and filtered to drivable road features using osmium. The resulting file is loaded with ox$graph_from_xml() to build a single drive-network graph covering the whole county. That graph is projected with project_graph(), and nearby nodes are merged into intersection points with consolidate_intersections().

```{r}
#| label: osmnx-setup

reticulate::conda_install("osmnx-env", packages = "osmium-tool", channel = "conda-forge")

# confirm it exists
file.exists(file.path(
  dirname(reticulate::conda_binary()), "..", "envs", "osmnx-env",
  "Library", "bin", "osmium.exe"
))

reticulate::use_condaenv("osmnx-env", required = TRUE)

# convert = FALSE: keep everything returned by these modules as live Python
# object references rather than letting reticulate auto-convert them to R
# the instant a call returns (see get_area_intersections() below for why).
ox <- reticulate::import("osmnx", convert = FALSE)
shapely_wkt <- reticulate::import("shapely.wkt", convert = FALSE)

ox$settings$use_cache <- TRUE
ox$settings$log_console <- TRUE
```

```{r}
library(sf)
cc_boundary_geo <- st_read("data/boundaries/cc_boundary.shp") |> st_transform(4326)
st_write(cc_boundary_geo, "data/boundaries/cc_boundary.geojson", delete_dsn = TRUE)

# if you don't already have the Illinois PBF downloaded somewhere, grab it:
download.file(
  "https://download.geofabrik.de/north-america/us/geofabrik_illinois-latest.osm.pbf",
  "illinois-latest.osm.pbf", mode = "wb", method = "libcurl"
)

reticulate::conda_run2(
  cmd = "osmium",
  args = c(
    "extract", "--polygon", "data/boundaries/cc_boundary.geojson",
    "data/osm/illinois-latest.osm.pbf",
    "-o", "data/osm/cook_extract.osm.pbf", "--overwrite"
  ),
  envname = "osmnx-env"
)

```

```{r}
#| label: filter-constants

highway_subtypes <- c(
  "primary", "primary_link", "unclassified", "residential",
  "secondary", "secondary_link", "service", "tertiary",
  "tertiary_link", "trunk", "trunk_link", "busway",
  "construction", "corridor"
)
service_exclusion <- c("parking_aisle", "driveway")

drive_filter <- glue(
  '["highway"~"{paste(highway_subtypes, collapse = "|")}"]',
  '["service"!~"{paste(service_exclusion, collapse = "|")}"]',
  '["access"!~"private"]["aeroway"!~".*"]["indoor"!~"yes"]'
)
```


```{r}
# 1. Keep only the drive-network highway subtypes (same list as dataprep.qmd's highway_subtypes)
reticulate::conda_run2(
  cmd = "osmium",
  args = c(
    "tags-filter", "data/osm/cook_extract.osm.pbf",
    "w/highway=primary,primary_link,unclassified,residential,secondary,secondary_link,service,tertiary,tertiary_link,trunk,trunk_link,busway,construction,corridor",
    "-o", "data/osm/cook_01_highway.osm.pbf", "--overwrite"
  ),
  envname = "osmnx-env"
)

# 2. Drop the service-road exclusions (parking aisles, driveways)
reticulate::conda_run2(
  cmd = "osmium",
  args = c(
    "tags-filter", "-i", "data/osm/cook_01_highway.osm.pbf",
    "w/service=parking_aisle,driveway",
    "-o", "data/osm/cook_02_service.osm.pbf", "--overwrite"
  ),
  envname = "osmnx-env"
)

# 3. Drop private-access ways
reticulate::conda_run2(
  cmd = "osmium",
  args = c(
    "tags-filter", "-i", "data/osm/cook_02_service.osm.pbf",
    "w/access=private",
    "-o", "data/osm/cook_03_access.osm.pbf", "--overwrite"
  ),
  envname = "osmnx-env"
)

# 4. Drop anything tagged as an aeroway (any value)
reticulate::conda_run2(
  cmd = "osmium",
  args = c(
    "tags-filter", "-i", "data/osm/cook_03_access.osm.pbf",
    "w/aeroway",
    "-o", "data/osm/cook_04_aeroway.osm.pbf", "--overwrite"
  ),
  envname = "osmnx-env"
)

# 5. Drop indoor=yes, and output straight to OSM XML (osmium's -f/--output-format
#    works on any command, so this folds the pbf-to-XML conversion into the last
#    filtering pass rather than needing a separate `osmium cat` step)
reticulate::conda_run2(
  cmd = "osmium",
  args = c(
    "tags-filter", "-i", "data/osm/cook_04_aeroway.osm.pbf",
    "w/indoor=yes",
    "-o", "data/osm/cook_drive_network.osm", "-f", "osm", "--overwrite"
  ),
  envname = "osmnx-env"
)

file.exists("data/osm/cook_drive_network.osm")
```

```{r}
G_drive <- ox$graph_from_xml("data/osm/cook_drive_network.osm", retain_all = FALSE)
G_drive <- ox$project_graph(G_drive)

intersections <- ox$consolidate_intersections(G_drive, rebuild_graph = FALSE, dead_ends = FALSE)

coords_df <- intersections$to_crs("EPSG:4326")$get_coordinates()
lon <- as.numeric(reticulate::py_to_r(coords_df$x$to_numpy()))
lat <- as.numeric(reticulate::py_to_r(coords_df$y$to_numpy()))

cook_intersections <- tibble::tibble(lon = lon, lat = lat)
```

```{r}

cook_intersections_geo <- cook_intersections %>% 
  st_as_sf(coords = c("lon", "lat"), crs = 4326) %>% 
  st_transform(crs = 3435)

study_area_intersections <- cook_intersections_geo %>% 
  st_join(study_area) %>% 
  drop_na() %>% 
  st_make_valid() %>% 
  st_as_sf() 

st_write(study_area_intersections, "data/intersections/study_area_intersections.gpkg", delete_dsn = TRUE)

```

### Sample of intersections
```{r}
library(dplyr)

all_intersections <- st_read("output/all_intersections.shp")

set.seed(42)

# add row number
all_intersections_id <- all_intersections %>% mutate(.id = row_number())

all_intersections_sample <- all_intersections_id %>%
  # group_by(region) %>%   # strata nested within geography
  slice_sample(n = 50)

test <- df %>% anti_join(train, by = ".id")

```

## Streets and Speed Limits

Street segments come from the Illinois OpenStreetMap extract, keeping only road features and clipping them to the county boundary. Each segment is assigned a speed limit from its posted value in OpenStreetMap where one is recorded. Where it isn't, a default is applied based on road type, such as 25 mph for residential streets, which matches the limit many Cook County suburbs post below the state's 30 mph urban default. A source field marks whether each segment's speed limit was posted or estimated, since the defaults are approximations rather than measured data.

```{r}
#| label: speed-limit-defaults

default_speed_by_highway <- c(
  motorway = 65, motorway_link = 45,
  trunk = 55, trunk_link = 40,
  primary = 45, primary_link = 35,
  secondary = 35, secondary_link = 30,
  tertiary = 30, tertiary_link = 25,
  unclassified = 30,
  residential = 25,
  living_street = 15,
  service = 15
)

# OSM tag values can themselves be semicolon-joined (e.g. "30;20") -- use
# the first for class-default lookup and mph parsing.
first_tag_value <- function(x) sub(";.*$", "", x)

parse_maxspeed_mph <- function(x) {
  first <- first_tag_value(x)
  # substr()/regexpr() only, not regmatches() -- regmatches() with a single
  # regexpr() result silently drops non-matching elements rather than
  # returning a same-length vector, which would misalign this against the
  # rest of the tibble whenever maxspeed is untagged (the common case).
  m <- regexpr("[0-9]+", first)
  starts <- as.integer(m)
  lens <- attr(m, "match.length")
  digits <- ifelse(starts > 0, substr(first, starts, starts + lens - 1L), NA_character_)
  suppressWarnings(as.numeric(digits))
}

resolve_speed_limit <- function(highway, maxspeed_raw) {
  tagged_mph <- parse_maxspeed_mph(maxspeed_raw)
  class_default <- unname(default_speed_by_highway[first_tag_value(highway)])
  tibble(
    maxspeed_mph = ifelse(!is.na(tagged_mph), tagged_mph, class_default),
    maxspeed_source = ifelse(!is.na(tagged_mph), "osm_tagged", "class_default")
  )
}
```

```{r}
#| label: streets-osm
#| # https://download.geofabrik.de/north-america/us/illinois.html

il_streets <- oe_read(
  "https://download.geofabrik.de/north-america/us/illinois-latest.osm.pbf",
  layer = "lines",
  extra_tags = c("maxspeed", "oneway"),
  query = "SELECT * FROM lines WHERE highway IS NOT NULL",
  quiet = TRUE
)

study_area_streets <- il_streets %>%
  filter(highway %in% highway_subtypes) %>%
  st_transform(3435) %>%
  st_make_valid() %>%
  st_intersection(st_geometry(cc_boundary)) %>%
  st_collection_extract("LINESTRING") %>%
  select(street_name = name, highway, maxspeed_raw = maxspeed, oneway) %>%
  mutate(resolve_speed_limit(highway, maxspeed_raw))

st_write(study_area_streets, "data/infrastructure/study_area_streets.gpkg", delete_dsn = TRUE)
```

## Walkability 
```{r}
walkability <- st_read("data/infrastructure/Walkability.shp") %>% 
  st_transform(crs=3435) %>%
  st_intersection(st_geometry(cc_boundary)) %>%
  st_as_sf() %>% 
  st_make_valid()

st_write(walkability, "data/infrastructure/walkability.gpkg", delete_dsn = TRUE)
```

## Land Use Inventory (CMAP 2020)

CMAP's 2020 Land Use Inventory (LUI) is downloaded from its ArcGIS feature service, and is clipped to Cook County. Features are first requested within Cook County's bounding box (server-side), which keeps the download from covering the full seven-county region, and then intersected with the county boundary.

```{r}
#| label: land-use-inventory

lui_url <- "https://services5.arcgis.com/LcMXE3TFhi1BSaCY/arcgis/rest/services/LUI20_geodatabase_v1_CMAP/FeatureServer"

# open the feature service; print `lui_fs` to list its layers if the layer
# id below needs to change
lui_fs <- arc_open(lui_url)
lui_layer <- get_layer(lui_fs, id = 0)

# arc_select() pages through the service's max record count automatically
lui_raw <- arc_select(
  lui_layer,
  filter_geom = st_as_sfc(st_bbox(cc_boundary))
)

land_use <- lui_raw %>%
  st_transform(3435) %>%
  st_make_valid() %>%
  st_intersection(st_geometry(cc_boundary)) %>%
  st_collection_extract("POLYGON") %>%
  st_as_sf()

st_write(land_use, "data/infrastructure/land_use_2020.gpkg", delete_dsn = TRUE)
```

## Annual Average Daily Traffic (AADT)
https://gis-idot.opendata.arcgis.com/datasets/annual-average-daily-traffic-aadt

```{r}
aadt <- st_read("data/infrastructure/aadt.geojson") %>% 
  st_transform(3435) %>%
  st_make_valid() %>%
  st_intersection(st_geometry(cc_boundary)) %>% 
  st_as_sf()

```



## Pedestrian Fatalities (FARS)

Fatal crash records for Cook County from 2010 to 2024 are pulled from NHTSA's Fatality Analysis Reporting System (FARS). The records are narrowed to crashes whose first harmful event involved a pedestrian. Crashes with missing or placeholder coordinates are dropped so that each remaining fatality can be mapped at its reported location. A readable date and time is added to each record for display on the map, using whichever date and time fields the data provides.

```{r}
#| label: crashes-fars

# Cook County, Illinois FIPS codes -- fixed constants for this project's
# single-county scope.
state_fips  <- 17L  # Illinois
county_fips <- 31L  # Cook County

crashes_fars <- read_csv(paste0(
  "https://crashviewer.nhtsa.dot.gov/crashviewer/CrashAPI/FARSData/GetFARSData?",
  "dataset=Accident&FromYear=2010&ToYear=2024&State=", state_fips,
  "&county=", county_fips, "&format=csv"
), show_col_types = FALSE) %>%
  filter(state == state_fips, county == county_fips)

ped_fatalities <- crashes_fars %>%
  filter(harm_evname == "Pedestrian") %>%
  filter(
    latitude  > -90,  latitude  < 90,
    longitud  > -180, longitud  < 180,
    !latitude  %in% c(77.7777, 88.8888, 99.9999),   # FARS sentinels for unknown/not-reported coordinates
    !longitud  %in% c(-77.7777, -88.8888, -99.9999)
  ) %>%
  st_as_sf(coords = c("longitud", "latitude"), crs = 4326)

# Build hover date/time text defensively -- check what's actually present
# in the CrashAPI response rather than assuming specific column names.
get_col_or_null <- function(df, col) if (col %in% names(df)) df[[col]] else NULL

monthname <- get_col_or_null(ped_fatalities, "monthname")
month_num <- get_col_or_null(ped_fatalities, "month")
day       <- get_col_or_null(ped_fatalities, "day")
year      <- get_col_or_null(ped_fatalities, "year")
hourname  <- get_col_or_null(ped_fatalities, "hourname")
hour_num  <- get_col_or_null(ped_fatalities, "hour")
minute    <- get_col_or_null(ped_fatalities, "minute")

fatality_date <- if (!is.null(monthname) && !is.null(day) && !is.null(year)) {
  paste(monthname, day, year)
} else if (!is.null(month_num) && !is.null(day) && !is.null(year)) {
  sprintf("%02d/%02d/%d", month_num, day, year)
} else {
  NA_character_
}

fatality_time <- if (!is.null(hourname)) {
  as.character(hourname)
} else if (!is.null(hour_num) && !is.null(minute)) {
  sprintf("%02d:%02d", hour_num, minute)
} else {
  NA_character_
}

ped_fatalities_formatted <- ped_fatalities %>%
  # Attach fatality_date/fatality_time as real columns before the join below,
  # rather than leaving them as loose chunk-local vectors -- otherwise
  # they never make it into the saved .gpkg (and, since drop_na(geoid) below
  # removes rows, couldn't safely be attached after the join anyway).
  mutate(fatality_date = fatality_date, fatality_time = fatality_time) %>%
  st_join(study_area) %>%
  drop_na(geoid) %>%
  st_make_valid() %>%
  st_as_sf()

st_write(ped_fatalities_formatted, "data/crashes/ped_fatalities.gpkg", delete_dsn = TRUE)
```

## Crash Records (IDOT)

Crash records for 2022 to 2024 are read from IDOT's annual crash files and combined into a single dataset. Each crash is mapped using its recorded coordinates, falling back to a second set of coordinates when the first is missing or falls outside Illinois. The date and time are built from the separate year, month, day and hour fields, since the combined date field can shift late-evening crashes to the next day. Each crash is flagged as a pedestrian or bicycle crash where applicable, and as a killed-or-seriously-injured (KSI) crash if anyone died or was seriously injured; a second output contains only the pedestrian and bicycle crashes.

```{r}
#| label: idot-crashes

# Required lowercased column names this function depends on -- checked
# explicitly below so a schema change in a future IDOT export (a renamed
# or dropped column) fails with a clear message naming the column, rather
# than a cryptic downstream "object not found" error from dplyr.
required_idot_cols <- c(
  "x", "y", "tscrashlongitude", "tscrashlatitude",
  "crashyr", "crashmonth", "crashday", "crashhour",
  "typeoffirstcrash", "totalfatals","totalinjured", "ainjuries"
)

read_crash_year <- function(year, dir) {
  file <- list.files(dir, pattern = paste0(year, ".*\\.csv$"),
                     full.names = TRUE, ignore.case = TRUE)
  if (length(file) == 0) {
    warning("No CSV found for ", year, " in ", dir, "; skipping.")
    return(NULL)
  }
  if (length(file) > 1) {
    warning("Multiple CSVs match ", year, "; using ", basename(file[1]))
    file <- file[1]
  }
  read_csv(file, col_types = cols(.default = col_character()),
           show_col_types = FALSE) %>%
    rename_with(tolower) %>%
    mutate(source_file = basename(file))
}

process_crashes <- function(years, dir, crs_out = 4326, ped_bike_only = FALSE) {
  crash_raw <- years %>%
    map(read_crash_year, dir = dir) %>%
    compact() %>%
    bind_rows() %>%
    type_convert(col_types = cols())

  missing_cols <- setdiff(required_idot_cols, names(crash_raw))
  if (length(missing_cols) > 0) {
    stop(
      "IDOT crash CSV is missing expected column(s): ",
      paste(missing_cols, collapse = ", "),
      ". Check the raw column headers (rename_with(tolower) is applied) ",
      "and update required_idot_cols/process_crashes() to match."
    )
  }

  in_il <- function(lon, lat) {
    !is.na(lon) & !is.na(lat) &
      between(lon, -91.6, -87.0) & between(lat, 36.9, 42.6)
  }

  # x/y are the ArcGIS-exported coordinate pair, distinct from
  # tscrashcoordinatex/y (a separate, likely projected pair not used here)
  # and from tscrashlongitude/tscrashlatitude (the fallback). Report how
  # often x/y actually resolves as in-Illinois lon/lat vs. falls back, as a
  # sanity check that x/y really is geographic and not itself projected.
  n_xy_valid <- sum(in_il(crash_raw$x, crash_raw$y), na.rm = TRUE)
  cli::cli_alert_info(
    "IDOT crashes: {n_xy_valid} of {nrow(crash_raw)} rows used x/y directly (rest fell back to TSCrashLongitude/TSCrashLatitude)."
  )

  crash_df <- crash_raw %>%
    mutate(
      lon = if_else(in_il(x, y), x, tscrashlongitude),
      lat = if_else(in_il(x, y), y, tscrashlatitude),
      # crash date/time built from crashyr/crashmonth/crashday/crashhour
      # rather than the raw crashdate field, which is UTC-shifted and can
      # move late-evening crashes to the next calendar day.
      crash_date = make_date(crashyr + 2000, crashmonth, crashday),
      crash_dth   = crash_date + hours(crashhour),
      ped_bike = case_when(
        typeoffirstcrash == "Pedestrian"   ~ "Pedestrian",
        typeoffirstcrash == "Pedalcyclist" ~ "Bicycle",
        TRUE                               ~ NA_character_
      ),
      ksi = totalfatals > 0 | ainjuries > 0
    ) %>%
    select(-any_of("crash_hour"))

  crash <- crash_df %>%
    filter(in_il(lon, lat)) %>%
    st_as_sf(coords = c("lon", "lat"), crs = 4326, remove = FALSE)

  if (ped_bike_only) crash <- filter(crash, !is.na(ped_bike))

  st_transform(crash, crs_out)
}

idot_years <- 2022:2024
idot_dir   <- "data/crashes"

il_crashes <- process_crashes(idot_years, dir = idot_dir)
il_ped_bike_crashes <- process_crashes(idot_years, dir = idot_dir, ped_bike_only = TRUE)

# join crashes with study area
cc_crashes <- il_crashes %>%
  select(
    x = lon,
    y = lat,
    year = crashyr,
    month = crashmonth,
    day = crashday,
    datetime = crash_dth,
    crash_type = collisiontypecode,
    ped_bike = typeoffirstcrash,
    fatals = totalfatals,
    injurs = totalinjured,
    ainjurs = ainjuries,
    binjurs = binjuries,
    cinjurs = cinjuries,
    ksi,
    severity = crashseverity,
  ) %>%
  st_transform(crs = 3435) %>%
  st_join(study_area) %>%
  drop_na(geoid) %>%
  st_make_valid() %>%
  st_as_sf() 

# join ped and bike crashes with study area
cc_ped_bike_crashes <- il_ped_bike_crashes %>%
  select(
    x = lon,
    y = lat,
    year = crashyr,
    month = crashmonth,
    day = crashday,
    datetime = crash_dth,
    crash_type = collisiontypecode,
    ped_bike,
    fatals = totalfatals,
    injurs = totalinjured,
    ainjurs = ainjuries,
    binjurs = binjuries,
    cinjurs = cinjuries,
    ksi,
    severity = crashseverity,
  ) %>% 
  st_transform(crs = 3435) %>% 
  st_join(study_area) %>% 
  drop_na(geoid) %>% 
  st_make_valid() %>% 
  st_as_sf() 

# write to folder
st_write(obj = cc_crashes, dsn = "data/crashes/cc_crashes_2022_2024.gpkg", append = FALSE)
st_write(cc_ped_bike_crashes, "data/crashes/cc_ped_bike_crashes_2022_2024.gpkg", append = FALSE)
```

## Area Deprivation Index (ADI)

This section produces the neighborhood disadvantage measures used in the Socioeconomic and Contextual Data table for Aim 2. It calculates the Area Deprivation Index (ADI), a composite score of neighborhood socioeconomic disadvantage based on income, education, employment, and housing, for two sets of areas: Chicago's community areas, built up from 2020 census tracts, and the municipalities of suburban Cook County. The underlying data come from the 2024 American Community Survey 5-year estimates, and the approach is adapted from an earlier analysis in adi_calculations.qmd.

The sociome R package usually handles ADI calculations automatically, but its standard function only works with built-in geographies such as states, counties, and census tracts. It can't produce scores for municipalities or for custom areas like Chicago community areas. To work around this, the code downloads the census data separately and passes it to the package's core ADI calculation step, a route the package's documentation supports. It requests the same census variables the package would use on its own, so the results follow the same method as standard ADI scores.

Unlike the national 1–100 percentiles published by the Neighborhood Atlas, these scores are relative to the areas included in the calculation. Higher values indicate greater disadvantage.

```{r}
#| label: adi-setup
# Set once per machine: tidycensus::census_api_key(Sys.getenv("CENSUS_API_KEY"), install = TRUE)
CENSUS_API_KEY <- Sys.getenv("CENSUS_API_KEY", unset = "8f6a0a83c8a2466e3e018a966846c86412d0bb6e")
tidycensus::census_api_key(CENSUS_API_KEY, install = TRUE, overwrite = TRUE)

adi_year   <- 2024
adi_survey <- "acs5"

# sociome's internal "set1" ACS variable list, reconstructed from
# data-raw/acs_vars.R (github.com/ClevelandClinicQHS/sociome) -- the set
# get_adi() uses for ACS5/ACS3/ACS1 data, year >= 2011 (excluding the
# 2015-2016-at-block-group exception, which doesn't apply here).
# calculate_adi() requires these exact column names (no "E"/"M" suffix), so
# the pull below uses output = "tidy" and pivots to wide manually rather
# than tidycensus::get_acs(output = "wide")'s "<var>E"/"<var>M" columns.
adi_acs_vars <- c(
  "B11005_001", "B15003_001", "B15003_002", "B15003_003", "B15003_004",
  "B15003_005", "B15003_006", "B15003_007", "B15003_008", "B15003_009",
  "B15003_010", "B15003_011", "B15003_012", "B15003_017", "B15003_018",
  "B15003_019", "B15003_020", "B15003_021", "B15003_022", "B15003_023",
  "B15003_024", "B15003_025", "B17010_001", "B17010_002", "B19001_002",
  "B19001_011", "B19001_012", "B19001_013", "B19001_014", "B19001_015",
  "B19001_016", "B19001_017", "B19113_001", "B11005_002", "B11005_005",
  "B23025_003", "B23025_005", "B25003_001", "B25003_002", "B25014_001",
  "B25014_005", "B25014_006", "B25014_007", "B25014_011", "B25014_012",
  "B25014_013", "B25044_001", "B25044_003", "B25044_010", "B25064_001",
  "B25077_001", "B25088_002", "C17002_001", "C17002_002", "C17002_003",
  "C17002_004", "C17002_005", "C24010_001", "C24010_003", "C24010_039"
)

# Four of the above are medians (dollar values), which can't be summed when
# aggregating tracts up to a community area -- a population-weighted mean of
# the tract medians approximates the aggregate median instead. Every other
# variable is a count, and gets summed.
adi_median_vars <- c("B19113_001", "B25064_001", "B25077_001", "B25088_002")
adi_sum_vars    <- setdiff(adi_acs_vars, adi_median_vars)

get_acs_tidy_wide <- function(..., variables) {
  tidycensus::get_acs(
    ...,
    variables = variables,
    year = adi_year,
    survey = adi_survey,
    output = "tidy"
  ) %>%
    select(GEOID, NAME, variable, estimate) %>%
    pivot_wider(names_from = variable, values_from = estimate)
}
```

### Chicago community area

This step calculates ADI for each of Chicago's 77 community areas. Census data aren't published by community area, so the code starts with 2024 American Community Survey estimates for every census tract in Cook County, using 2020 tract boundaries. It assigns each tract to the community area it falls within, using the tract boundary file from the City of Chicago Data Portal and the community area boundaries defined earlier in the Study Area Geography section. It then combines the tract data within each community area and calculates one ADI score per area.

```{r}
#| label: adi-community-areas

tracts_raw <- get_acs_tidy_wide(
  geography = "tract",
  state = "17",
  county = "031",
  variables = c(adi_acs_vars, "B01003_001")  # + total population, to weight the medians below
)

tract_boundaries <- st_read(
  "https://data.cityofchicago.org/resource/7qjk-677h.geojson",
  quiet = TRUE
) %>%
  st_transform(3435)

tract_geoid_col <- intersect(
  names(tract_boundaries), c("geoid20", "GEOID20", "geoid", "GEOID")
)[1]

tract_ca_lookup <- tract_boundaries %>%
  mutate(GEOID = tract_full) %>%
  st_point_on_surface() %>%
  st_join(community_areas %>% select(commarea_geoid = geoid, commarea_name = name)) %>%
  st_drop_geometry() %>%
  select(GEOID, commarea_geoid, commarea_name) %>%
  filter(!is.na(commarea_geoid))  # tract centroids outside any community area

tracts_with_ca <- tracts_raw %>%
  mutate(GEOID = as.character(GEOID)) %>%
  inner_join(tract_ca_lookup, by = "GEOID")

adi_ca_agg <- tracts_with_ca %>%
  group_by(commarea_geoid, commarea_name) %>%
  summarise(
    across(all_of(adi_sum_vars), \(x) sum(x, na.rm = TRUE)),
    across(all_of(adi_median_vars), \(x) weighted.mean(x, w = B01003_001, na.rm = TRUE)),
    .groups = "drop"
  ) %>%
  rename(GEOID = commarea_geoid, NAME = commarea_name) %>%
  relocate(GEOID, NAME)

adi_community_areas <- calculate_adi(adi_ca_agg, keep_indicators = TRUE)

write_csv(adi_community_areas, "data/census/adi_community_areas.csv")
```

### Suburban municipality

This step calculates ADI for each municipality in suburban Cook County. The Census Bureau publishes data for cities, villages, and towns as "places," but it can only provide them for an entire state at once, not by county. The code therefore downloads 2024 American Community Survey data for all of Illinois's roughly 1,300 places and keeps only the Cook County municipalities identified in the Study Area Geography section. Filtering before the calculation matters because ADI scores are relative to the set of areas they are computed from. Each suburb is therefore scored against other suburban Cook County municipalities, not against communities statewide.

```{r}
#| label: adi-places

places_raw <- get_acs_tidy_wide(
  geography = "place",
  state = "17",
  variables = adi_acs_vars
)

suburban_place_geoids <- places_cook_county$geoid

places_cc_raw <- places_raw %>%
  mutate(GEOID = as.character(GEOID)) %>%
  filter(GEOID %in% suburban_place_geoids)

# places_cook_county$geoid's exact provenance (Metopio's place topojson,
# Study Area Geography above) wasn't independently verified as matching
# tidycensus's Census Place GEOIDs from this environment -- if this warns,
# check that assumption before trusting adi_places.
if (nrow(places_cc_raw) < length(suburban_place_geoids) * 0.9) {
  cli::cli_alert_warning(
    "Matched only {nrow(places_cc_raw)} of {length(suburban_place_geoids)} suburban municipalities to Census place GEOIDs."
  )
}

adi_places <- calculate_adi(places_cc_raw, keep_indicators = TRUE)

write_csv(adi_places, "data/census/adi_suburban_places.csv")
```

### Study Area ADI
```{r}
adi_study_area <- adi_community_areas %>% 
  bind_rows(adi_places)

write_csv(adi_study_area, "data/census/adi_study_area.csv")
```

::: {.callout-note collapse="true"}
## Note: two separate reference areas
ADI is a relative measure -- its value for a given area depends on which other areas were included in the same `calculate_adi()` call (its "reference area"). The two calculations above use separate reference areas (77 community areas; suburban municipalities), so `adi_community_areas$ADI` and `adi_places$ADI` are not on a directly comparable scale, even though both used the same ACS variables/year. Combining them into one reference area (one `calculate_adi()` call over all ~200 units) would make them comparable, at the cost of mixing a tract-aggregated construct with a directly-measured one -- worth resolving explicitly in `phase02.qmd` before Aim 2 analysis, rather than assumed away here.
:::

## Read, Format and Save

### Read
```{r}

county_boundary <- st_read(dsn = "data/boundaries/cc_boundary.gpkg") %>% st_transform(crs=4326)
cog_boundaries <- st_read(dsn = "data/boundaries/cogs_boundaries.gpkg") %>% st_transform(crs=4326)
study_area <- st_read(dsn = "data/boundaries/study_area.gpkg") %>% st_transform(crs=4326)
intersections <- st_read("data/intersections/study_area_intersections.gpkg") %>% st_transform(crs=4326)
streets <- st_read("data/infrastructure/study_area_streets.gpkg") %>% st_transform(crs=4326)
walkability <- st_read("data/infrastructure/walkability.gpkg") %>% st_transform(crs=4326)
ped_fatalities <- st_read("data/crashes/ped_fatalities.gpkg") %>% st_transform(crs=4326)
ped_bike_crashes <- st_read("data/crashes/cc_ped_bike_crashes_2022_2024.gpkg") %>% st_transform(crs=4326)
adi <- read.csv("data/census/adi_study_area.csv")

```


Save the assembled `sf`/`tibble` tables to a single `.RData` file. `phase01.qmd` loads this file to build an interactive data-exploration map; downstream analysis in `phase02.qmd` will read from these same tables (or their saved shapefiles/CSVs under `output/`).

```{r}
#| label: save-dataprep

save(
  study_area,
  county_boundary,
  cog_boundaries,
  streets,
  intersections,
  walkability,
  ped_fatalities,
  ped_bike_crashes,
  adi,
  file = "data/phase01_dataprep.RData"
)
```
