---
execute:
  eval: false
  warning: false
editor_options:
  chunk_output_type: console
---

# Data Preparation Code {#sec-dataprep}

## Purpose

This chapter contains the code that builds the `sf` data tables that support [Dataset Development](phase01.html)'s dataset-development tasks: the countywide study-area geography, street intersections, OSM crossing edges and the crossing-distance comparison sample, the street/speed-limit network, pedestrian fatality records (NHTSA FARS), and crash records (IDOT). It ends by saving the resulting tables to `data/phase01_dataprep.RData`, which [Dataset Development](phase01.html) loads (a subset of) to build an interactive map for data exploration.

## Activate R Libraries

```{r}
#| label: r-packages
#| eval: true

library(sf)            # simple features: read/write, transform and join spatial data
library(reticulate)    # call Python (osmnx, shapely) from R for intersection extraction
library(dplyr)         # data manipulation verbs (filter, mutate, select, joins)
library(tidyr)         # reshaping and missing-value helpers (drop_na, pivot_*)
library(purrr)         # functional iteration (map) over crash files by year
library(readr)         # fast CSV reading/writing (IDOT crashes, ADI tables)
library(glue)          # string interpolation (osmium filter expressions)
library(cli)           # console alerts/warnings for data-quality checks
library(tigris)        # download Census TIGER/Line boundaries (Cook County)
library(units)         # unit-aware areas and lengths (m^2 -> mi^2)
library(osmextract)    # read OpenStreetMap extracts (street network, speed limits)
library(lubridate)     # date/time construction for crash records
library(stringr)       # string helpers (str_replace, str_starts)
library(tidycensus)    # download American Community Survey (ACS) data
library(sociome)       # compute Area Deprivation Index (ADI) from ACS data
library(arcgislayers)  # query ArcGIS feature services (CMAP land use)
```

## Study Area Geography

Combines the City of Chicago's 77 community areas with suburban Cook County's incorporated municipalities (≥4 sq mi) into a single, internally consistent geography (`study_area`, N ≈ 200), identified by `geoid` and `name`, with a `type` field distinguishing the two as either `community area` or `municipality`.

### Cook County boundary
```{r}
#| label: study-area

# download US Census TIGER/Line Files using tigris package
counties_il <- counties(state = "17", cb = TRUE)

cc_boundary <- counties_il %>%
  filter(GEOID == "17031") %>%
  st_transform(3435) %>%
  select(geoid_county = GEOID, name_county = NAME,
         area_land = ALAND, area_water = AWATER) %>%
  mutate(
    area_land_sqmi  = set_units(set_units(area_land, m^2), mi^2),
    area_water_sqmi = set_units(set_units(area_water, m^2), mi^2),
    area_sqmi = area_land_sqmi + area_water_sqmi
  ) %>%
  select(-area_land, -area_water)

st_write(cc_boundary, "data/boundaries/cc_boundary.gpkg", delete_dsn = TRUE)
```

### Study areas boundaries
```{r}
# download community areas from Chicago Health Atlas
community_areas <- st_read(
  "https://metopio.blob.core.windows.net/lalage/atlas/3/shapes/communityareas.topo.json",
  quiet = TRUE
) %>%
  st_as_sf() %>%
  st_set_crs(4326) %>%
  st_transform(3435) %>%
  rename(geoid = id) %>%
  mutate(
    geoid = str_replace(geoid, "1714000-", ""),
    location = "Chicago, IL",
    type = "community area"
  )

# download municipalities from Cook County Health Atlas
places_cook_county <- st_read(
  "https://metopio.blob.core.windows.net/lalage/atlas/6/shapes/SCC-place.topo.json",
  quiet = TRUE
) %>%
  st_as_sf() %>%
  st_set_crs(4326) %>%
  st_transform(3435) %>%
  rename(geoid = id) %>%
  mutate(
    sqmi = as.numeric(st_area(geometry)) * 0.0000003861,
    location = "suburban Cook County",
    type = "municipality"
  ) %>%
  filter(sqmi >= 4) %>%
  select(-sqmi)

study_area <- places_cook_county %>%
  bind_rows(community_areas) %>%
  st_as_sf() %>%
  st_make_valid()

st_write(study_area, "data/boundaries/study_area.gpkg", delete_dsn = TRUE)

```

## Councils of mayors boundaries (CMAP)
```{r}
# import, process councils of Mayors boundaries for Cook County
cogs_boundaries <- st_read(dsn="data/boundaries/councils_of_mayors.shp") %>% 
  rename_with(tolower) %>% 
  select(council,
         name = fullname) %>% 
  st_transform(crs = 3435) 

st_write(cogs_boundaries, "data/boundaries/cogs_boundaries.gpkg", delete_dsn = TRUE)
```

## Land Use Inventory (CMAP 2020)

CMAP's 2020 Land Use Inventory (LUI) is downloaded from its ArcGIS feature service, and is clipped to Cook County. Features are first requested within Cook County's bounding box (server-side), which keeps the download from covering the full seven-county region, and then intersected with the county boundary. https://cmap.illinois.gov/data/land-use/land-use-inventory/

```{r}
#| label: land-use-inventory

lui_url <- "https://services5.arcgis.com/LcMXE3TFhi1BSaCY/arcgis/rest/services/LUI20_geodatabase_v1_CMAP/FeatureServer"

# open the feature service; print `lui_fs` to list its layers if the layer
# id below needs to change
lui_fs <- arc_open(lui_url)
lui_layer <- get_layer(lui_fs, id = 0)

# arc_select() pages through the service's max record count automatically
lui_raw <- arc_select(
  lui_layer)

study_area_land_use <- lui_raw %>%
  st_transform(3435) %>%
  st_make_valid() %>%
  st_intersection(st_geometry(cc_boundary)) %>%
  st_collection_extract("POLYGON") %>%
  st_as_sf()

study_area_land_use_simplified <- study_area_land_use %>%
  mutate(
    lu_code = as.character(LANDUSE),
    lu_category = case_when(
      str_starts(lu_code, "11") ~ "Residential",
      str_starts(lu_code, "12") ~ "Commercial",
      str_starts(lu_code, "13") ~ "Institutional",
      str_starts(lu_code, "14") ~ "Industrial",
      str_starts(lu_code, "15") ~ "Transportation & Utilities",
      str_starts(lu_code, "2")  ~ "Agriculture",
      str_starts(lu_code, "3")  ~ "Open Space",
      str_starts(lu_code, "4")  ~ "Vacant",
      str_starts(lu_code, "5")  ~ "Water",
      TRUE                      ~ "Other / Not Classified"
    )
  ) %>%
  group_by(lu_category) %>%
  summarise(.groups = "drop") %>%
  st_simplify(preserveTopology = TRUE, dTolerance = 10)

st_write(study_area_land_use, "data/boundaries/land_use_2020.gpkg", delete_dsn = TRUE)
st_write(study_area_land_use_simplified, "data/boundaries/land_use_simplified_2020.gpkg", delete_dsn = TRUE)
```

## Walkability (CMAP)
```{r}
walkability <- st_read("data/infrastructure/Walkability.shp") %>% 
  st_transform(crs=3435) %>%
  st_intersection(st_geometry(cc_boundary)) %>%
  st_as_sf() %>% 
  st_make_valid()

st_write(walkability, "data/infrastructure/walkability.gpkg", delete_dsn = TRUE)
```

## Trip Generation Zones (CMAP)
https://datahub.cmap.illinois.gov/datasets/d52c6a01a673495698e57da02d33d4fe_0/explore?location=41.912110%2C-87.928194%2C8


## Travel Demand Model Data (CMAP)
https://datahub.cmap.illinois.gov/documents/a14d641dbe184704ab597479b103986c/explore


## Street Intersections (OSM)

Street intersections are derived offline from an Illinois-wide OpenStreetMap extract, downloaded from [GeoFabrik](https://download.geofabrik.de). The extract is first clipped to the Cook County and filtered to drivable road features using osmium. The resulting file is loaded with ox$graph_from_xml() to build a single drive-network graph covering the whole county. That graph is projected with project_graph(), and nearby nodes are merged into intersection points with consolidate_intersections().

```{r}
#| label: osmnx-setup

reticulate::conda_install("osmnx-env", packages = "osmium-tool", channel = "conda-forge")

# confirm it exists
file.exists(file.path(
  dirname(reticulate::conda_binary()), "..", "envs", "osmnx-env",
  "Library", "bin", "osmium.exe"
))

reticulate::use_condaenv("osmnx-env", required = TRUE)

# convert = FALSE: keep everything returned by these modules as live Python
# object references rather than letting reticulate auto-convert them to R
# the instant a call returns (see get_area_intersections() below for why).
ox <- reticulate::import("osmnx", convert = FALSE)
shapely_wkt <- reticulate::import("shapely.wkt", convert = FALSE)

ox$settings$use_cache <- TRUE
ox$settings$log_console <- TRUE
```

```{r}
library(sf)
cc_boundary_geo <- st_read("data/boundaries/cc_boundary.shp") |> st_transform(4326)
st_write(cc_boundary_geo, "data/boundaries/cc_boundary.geojson", delete_dsn = TRUE)

# if you don't already have the Illinois PBF downloaded somewhere, grab it:
download.file(
  "https://download.geofabrik.de/north-america/us/geofabrik_illinois-latest.osm.pbf",
  "illinois-latest.osm.pbf", mode = "wb", method = "libcurl"
)

reticulate::conda_run2(
  cmd = "osmium",
  args = c(
    "extract", "--polygon", "data/boundaries/cc_boundary.geojson",
    "data/osm/illinois-latest.osm.pbf",
    "-o", "data/osm/cook_extract.osm.pbf", "--overwrite"
  ),
  envname = "osmnx-env"
)

```

```{r}
#| label: filter-constants

highway_subtypes <- c(
  "primary", "primary_link", "unclassified", "residential",
  "secondary", "secondary_link", "service", "tertiary",
  "tertiary_link", "trunk", "trunk_link", "busway",
  "construction", "corridor"
)
service_exclusion <- c("parking_aisle", "driveway")

drive_filter <- glue(
  '["highway"~"{paste(highway_subtypes, collapse = "|")}"]',
  '["service"!~"{paste(service_exclusion, collapse = "|")}"]',
  '["access"!~"private"]["aeroway"!~".*"]["indoor"!~"yes"]'
)
```


```{r}
# 1. Keep only the drive-network highway subtypes (same list as dataprep.qmd's highway_subtypes)
reticulate::conda_run2(
  cmd = "osmium",
  args = c(
    "tags-filter", "data/osm/cook_extract.osm.pbf",
    "w/highway=primary,primary_link,unclassified,residential,secondary,secondary_link,service,tertiary,tertiary_link,trunk,trunk_link,busway,construction,corridor",
    "-o", "data/osm/cook_01_highway.osm.pbf", "--overwrite"
  ),
  envname = "osmnx-env"
)

# 2. Drop the service-road exclusions (parking aisles, driveways)
reticulate::conda_run2(
  cmd = "osmium",
  args = c(
    "tags-filter", "-i", "data/osm/cook_01_highway.osm.pbf",
    "w/service=parking_aisle,driveway",
    "-o", "data/osm/cook_02_service.osm.pbf", "--overwrite"
  ),
  envname = "osmnx-env"
)

# 3. Drop private-access ways
reticulate::conda_run2(
  cmd = "osmium",
  args = c(
    "tags-filter", "-i", "data/osm/cook_02_service.osm.pbf",
    "w/access=private",
    "-o", "data/osm/cook_03_access.osm.pbf", "--overwrite"
  ),
  envname = "osmnx-env"
)

# 4. Drop anything tagged as an aeroway (any value)
reticulate::conda_run2(
  cmd = "osmium",
  args = c(
    "tags-filter", "-i", "data/osm/cook_03_access.osm.pbf",
    "w/aeroway",
    "-o", "data/osm/cook_04_aeroway.osm.pbf", "--overwrite"
  ),
  envname = "osmnx-env"
)

# 5. Drop indoor=yes, and output straight to OSM XML (osmium's -f/--output-format
#    works on any command, so this folds the pbf-to-XML conversion into the last
#    filtering pass rather than needing a separate `osmium cat` step)
reticulate::conda_run2(
  cmd = "osmium",
  args = c(
    "tags-filter", "-i", "data/osm/cook_04_aeroway.osm.pbf",
    "w/indoor=yes",
    "-o", "data/osm/cook_drive_network.osm", "-f", "osm", "--overwrite"
  ),
  envname = "osmnx-env"
)

file.exists("data/osm/cook_drive_network.osm")
```

```{r}
G_drive <- ox$graph_from_xml("data/osm/cook_drive_network.osm", retain_all = FALSE)
G_drive <- ox$project_graph(G_drive)

intersections <- ox$consolidate_intersections(G_drive, rebuild_graph = FALSE, dead_ends = FALSE)

coords_df <- intersections$to_crs("EPSG:4326")$get_coordinates()
lon <- as.numeric(reticulate::py_to_r(coords_df$x$to_numpy()))
lat <- as.numeric(reticulate::py_to_r(coords_df$y$to_numpy()))

cook_intersections <- tibble::tibble(lon = lon, lat = lat)
```

```{r}

cook_intersections_geo <- cook_intersections %>% 
  st_as_sf(coords = c("lon", "lat"), crs = 4326) %>% 
  st_transform(crs = 3435)

study_area_intersections <- cook_intersections_geo %>% 
  st_join(study_area) %>% 
  drop_na() %>% 
  st_make_valid() %>% 
  st_as_sf() 

st_write(study_area_intersections, "data/intersections/study_area_intersections.gpkg", delete_dsn = TRUE)

```

## OSM crossing edges

The grow-cut step of the reference methodology measures each crossing along an OSM crossing line, so the comparison sample (and later the countywide run) is limited to intersections with an OSM crossing nearby. The reference code builds these lines in `osm_ingest_local()` (`inference/osm_ingest.py` in [agupta01/crossing-distances](https://github.com/agupta01/crossing-distances)) from three Overpass queries for the whole county:

1. the full, unsimplified OSM network (`network_type = "all"`), from which every edge that is not a pedestrian crossing is removed;
2. the drive network, from which it derives intersections;
3. crossing features (`highway=crossing` or any `crossing` tag).

Here the same function runs unchanged, but the three queries read local OSM XML files cut from the Cook County extract above. This takes minutes instead of hours of Overpass queries, and uses the same OSM snapshot as the intersections. The drive network is `cook_drive_network.osm` from the previous step: its five osmium passes apply the same tag filter as the reference code's drive query (`drive_filter`). The other two files are built here. The full-network pass mirrors OSMnx's own `"all"` filter: ways with a `highway` tag, excluding areas and highways that are abandoned, planned, proposed, razed, raceways, platforms, rest areas or services.

```{r}
#| label: osm-crossing-files

# 1. Full network: every highway way, minus the classes OSMnx's "all" filter drops
reticulate::conda_run2(
  cmd = "osmium",
  args = c(
    "tags-filter", "data/osm/cook_extract.osm.pbf",
    "w/highway",
    "-o", "data/osm/cook_all_01_highway.osm.pbf", "--overwrite"
  ),
  envname = "osmnx-env"
)

reticulate::conda_run2(
  cmd = "osmium",
  args = c(
    "tags-filter", "-i", "data/osm/cook_all_01_highway.osm.pbf",
    "w/area=yes",
    "w/highway=abandoned,construction,no,planned,platform,proposed,raceway,razed,rest_area,services",
    "-o", "data/osm/cook_all_network.osm", "-f", "osm", "--overwrite"
  ),
  envname = "osmnx-env"
)

# 2. Crossing features: nodes, ways and relations tagged highway=crossing or
#    carrying any crossing tag (osmium adds the nodes that crossing ways reference)
reticulate::conda_run2(
  cmd = "osmium",
  args = c(
    "tags-filter", "data/osm/cook_extract.osm.pbf",
    "nwr/highway=crossing", "nwr/crossing",
    "-o", "data/osm/cook_crossings.osm", "-f", "osm", "--overwrite"
  ),
  envname = "osmnx-env"
)

file.exists(c("data/osm/cook_all_network.osm", "data/osm/cook_crossings.osm",
              "data/osm/cook_drive_network.osm"))
```

The reference repo is cloned next to the OSM files, and the packages its OSM code imports are added to `osmnx-env`. The chunk also binds reticulate to `osmnx-env`, so this section can be run on its own in a fresh R session. Run it before the Python chunk below. `modal` is needed only because the reference module imports it at load time; nothing is sent to Modal.

```{r}
#| label: reference-repo

reticulate::py_config()$python             # ...r-miniconda/envs/osmnx-env/python.exe
reticulate::py_module_available("osmnx")   # TRUE

# Bind reticulate to osmnx-env before any Python runs. reticulate can bind to
# only one Python per R session: if a Python chunk runs first (e.g. after an
# R restart), it starts reticulate's default Python, which has no osmnx, and
# `import osmnx` fails until R is restarted.
reticulate::use_condaenv("osmnx-env", required = TRUE)
if (!reticulate::py_module_available("osmnx")) {
  stop("Python is bound to ", reticulate::py_config()$python, ", which has no osmnx. ",
       "This needs the conda env osmnx-env (not a virtualenv of the same name). ",
       "RStudio binds its own default Python when a Python chunk runs first: set ",
       "Tools > Project Options > Python to the conda env osmnx-env, restart R, ",
       "and run this chunk before the Python chunk below.")
}

if (!dir.exists("data/osm/crossing-distances")) {
  system2("git", c("clone", "https://github.com/agupta01/crossing-distances.git",
                   "data/osm/crossing-distances"))
}
reticulate::conda_install("osmnx-env", c("modal", "geopy", "tqdm"), pip = TRUE)
```

`osm_ingest_local()` is then called with OSMnx's two place-based download functions pointed at the local files. The reference code expects the OSMnx 1.x feature index names (`element_type`, `osmid`); OSMnx 2.x renamed them to `element` and `id`, so the replacement for `features_from_place` renames them back. Building the unsimplified countywide graph is the memory-heavy step (plan on 16–32 GB of RAM).

```{python}
#| label: osm-crossing-edges

import os
import sys

sys.path.insert(0, os.path.abspath("data/osm/crossing-distances"))
os.environ.setdefault("MODAL_ENVIRONMENT", "cook")  # read when the reference module loads

import osmnx as ox
from inference import osm_ingest

ALL_XML = "data/osm/cook_all_network.osm"
DRIVE_XML = "data/osm/cook_drive_network.osm"
CROSSING_XML = "data/osm/cook_crossings.osm"


def graph_from_local(query, network_type="all", simplify=True, retain_all=False,
                     custom_filter=None, **kwargs):
    # the reference code passes a custom_filter only for the drive network
    path = DRIVE_XML if custom_filter else ALL_XML
    return ox.graph_from_xml(path, simplify=simplify, retain_all=retain_all)


def features_from_local(query, tags):
    gdf = ox.features_from_xml(CROSSING_XML, tags=tags)
    return gdf.rename_axis(index={"element": "element_type", "id": "osmid"})

ox.graph_from_place = graph_from_local
ox.features_from_place = features_from_local

osm_ingest.osm_ingest_local(place="Cook County, Illinois, USA", filepath="data/osm")
```

The function writes `crosswalk_edges_v2_0.shp` (crossing lines, EPSG:3857), `raw_intersection_coordinates_v2_0.csv` (its own drive-network intersections, which duplicate `all_intersections`) and `intersection_coordinates_v2_0.csv` to `data/osm/`. Copy the `crosswalk_edges_v2_0.*` files to `~/crosswalk/data/osm/` on the server in place of the runbook's step 4.1.

```{r}
#| label: osm-crossing-edges-check

crosswalk_edges <- st_read("data/osm/crosswalk_edges_v2_0.shp") %>%
  st_transform(crs = 3435)

nrow(crosswalk_edges)  # expect tens of thousands of lines
```

### Candidate pool

Equivalent to `01_build_candidates.py` in [Appendix B: Pipeline Runbook](phase01_modal_runbook.html) (step 4.2). It keeps intersections with an OSM crossing edge within 25 m, labels each as Chicago or suburban, and computes its position within the zoom-19 Google tile grid (`tile_offset`: 0 = tile centre, 1 = tile edge). It then draws up to 300 per geography. The `id` follows the reference repo's clip naming (e.g. `41878123N_87629456W`), which the Modal fetch step also uses, so later tables join on it. The draws use R's random number generator, so they differ from the Python script's at the same seed.

```{r}
#| label: candidate-pool

radius_ft      <- 25 * 3.28084  # 25 m in EPSG:3435 US feet
pool_per_group <- 300

crosswalk_id <- function(lat, lon) {
  sprintf("%.0f%s_%.0f%s",
          round(abs(lat) * 1e6), if_else(lat > 0, "N", "S"),
          round(abs(lon) * 1e6), if_else(lon > 0, "E", "W"))
}

# position within the Web Mercator tile containing the point
tile_offset <- function(lat, lon, z = 19) {
  n  <- 2^z
  xt <- (lon + 180) / 360 * n
  yt <- (1 - asinh(tan(lat * pi / 180)) / pi) / 2 * n
  2 * pmax(abs(xt %% 1 - 0.5), abs(yt %% 1 - 0.5))
}

all_intersections <- st_read("data/intersections/study_area_intersections.gpkg") %>%
  st_transform(crs = 3435)

nearest_edge <- st_nearest_feature(all_intersections, crosswalk_edges)

candidates <- all_intersections %>%
  mutate(dist_to_osm_crossing_ft = as.numeric(
    st_distance(all_intersections, crosswalk_edges[nearest_edge, ], by_element = TRUE)
  )) %>%
  filter(dist_to_osm_crossing_ft <= radius_ft)

cli::cli_alert_info(
  "{nrow(candidates)} of {nrow(all_intersections)} intersections have an OSM crossing within 25 m"
)

ll <- st_coordinates(st_transform(candidates, 4326))

set.seed(42)
candidates <- candidates %>%
  st_drop_geometry() %>%
  mutate(
    lat = round(ll[, "Y"], 6),
    lon = round(ll[, "X"], 6),
    id = crosswalk_id(lat, lon),
    geography = if_else(location == "Chicago, IL", "chicago", "suburban"),
    tile_offset = tile_offset(lat, lon)
  ) %>%
  distinct(id, .keep_all = TRUE) %>%
  group_by(geography) %>%
  slice_sample(n = pool_per_group) %>%
  ungroup() %>%
  select(id, lat, lon, geography, name, tile_offset, dist_to_osm_crossing_ft)

count(candidates, geography)  # expect 300 per geography

dir.create("data/sample", showWarnings = FALSE)
write_csv(candidates, "data/sample/candidates.csv")
```

Copy `data/sample/candidates.csv` to `~/crosswalk/data/sample/` on the server and run the canopy screen (runbook step 4.3). It writes `pool_ndvi.csv`, with `veg_frac`, the share of each 4-band clip's pixels with NDVI above 0.2. Copy it back to `data/sample/`.

### Evaluation sample and pilot set

Equivalent to the runbook's `02_select_sample.py` (step 4.4). Candidates are split into canopy terciles by `veg_frac` and into tile centre/edge at the median `tile_offset`. The middle canopy tercile is dropped so the high/low contrast is clean, leaving 8 strata (geography × canopy × tile position). Each stratum contributes `ceiling(n / 8)` intersections, 19 for `n = 150`, or 152 in total. The pilot set draws 15 more per geography from the remaining candidates, middle tercile included, and never overlaps the evaluation sample. It is used only to check segmentation settings on county imagery (runbook step 4.8).

```{r}
#| label: evaluation-sample

n_eval  <- 150
n_pilot <- 30

pool_ndvi <- read_csv("data/sample/pool_ndvi.csv")
if ("error" %in% names(pool_ndvi)) pool_ndvi <- filter(pool_ndvi, is.na(error))

pool <- candidates %>%
  inner_join(select(pool_ndvi, id, veg_frac, ndvi_mean), by = "id")

canopy_breaks <- quantile(pool$veg_frac, c(1/3, 2/3))

pool <- pool %>%
  mutate(
    canopy = cut(veg_frac, c(-1, canopy_breaks, 2), labels = c("low", "mid", "high")),
    tile_pos = if_else(tile_offset >= median(tile_offset), "edge", "centre")
  )

strata <- filter(pool, canopy != "mid")
per_cell <- ceiling(n_eval / n_distinct(strata$geography, strata$canopy, strata$tile_pos))

set.seed(42)
eval_sample <- strata %>%
  group_by(geography, canopy, tile_pos) %>%
  slice_sample(n = per_cell) %>%
  ungroup()

count(eval_sample, geography, canopy, tile_pos)  # thin cells: raise pool_per_group to 600

set.seed(43)
pilot_sample <- pool %>%
  anti_join(eval_sample, by = "id") %>%
  group_by(geography) %>%
  slice_sample(n = ceiling(n_pilot / 2)) %>%
  ungroup()

write_csv(eval_sample, "data/sample/sample.csv")
write_csv(pilot_sample, "data/sample/pilot.csv")
```

Copy `sample.csv` and `pilot.csv` to `~/crosswalk/data/sample/` on the server and continue with the runbook at step 4.5.

## Streets and Speed Limits (OSM)

Street segments come from the Illinois OpenStreetMap extract, keeping only road features and clipping them to the county boundary. Each segment is assigned a speed limit from its posted value in OpenStreetMap where one is recorded. Where it isn't, a default is applied based on road type, such as 25 mph for residential streets, which matches the limit many Cook County suburbs post below the state's 30 mph urban default. A source field marks whether each segment's speed limit was posted or estimated, since the defaults are approximations rather than measured data.

```{r}
#| label: speed-limit-defaults

default_speed_by_highway <- c(
  motorway = 65, motorway_link = 45,
  trunk = 55, trunk_link = 40,
  primary = 45, primary_link = 35,
  secondary = 35, secondary_link = 30,
  tertiary = 30, tertiary_link = 25,
  unclassified = 30,
  residential = 25,
  living_street = 15,
  service = 15
)

# OSM tag values can themselves be semicolon-joined (e.g. "30;20") -- use
# the first for class-default lookup and mph parsing.
first_tag_value <- function(x) sub(";.*$", "", x)

parse_maxspeed_mph <- function(x) {
  first <- first_tag_value(x)
  # substr()/regexpr() only, not regmatches() -- regmatches() with a single
  # regexpr() result silently drops non-matching elements rather than
  # returning a same-length vector, which would misalign this against the
  # rest of the tibble whenever maxspeed is untagged (the common case).
  m <- regexpr("[0-9]+", first)
  starts <- as.integer(m)
  lens <- attr(m, "match.length")
  digits <- ifelse(starts > 0, substr(first, starts, starts + lens - 1L), NA_character_)
  suppressWarnings(as.numeric(digits))
}

resolve_speed_limit <- function(highway, maxspeed_raw) {
  tagged_mph <- parse_maxspeed_mph(maxspeed_raw)
  class_default <- unname(default_speed_by_highway[first_tag_value(highway)])
  tibble(
    maxspeed_mph = ifelse(!is.na(tagged_mph), tagged_mph, class_default),
    maxspeed_source = ifelse(!is.na(tagged_mph), "osm_tagged", "class_default")
  )
}
```

```{r}
#| label: streets-osm
#| # https://download.geofabrik.de/north-america/us/illinois.html

il_streets <- oe_read(
  "https://download.geofabrik.de/north-america/us/illinois-latest.osm.pbf",
  layer = "lines",
  extra_tags = c("maxspeed", "oneway"),
  query = "SELECT * FROM lines WHERE highway IS NOT NULL",
  quiet = TRUE
)

study_area_streets <- il_streets %>%
  filter(highway %in% highway_subtypes) %>%
  st_transform(3435) %>%
  st_make_valid() %>%
  st_intersection(st_geometry(cc_boundary)) %>%
  st_collection_extract("LINESTRING") %>%
  select(street_name = name, highway, maxspeed_raw = maxspeed, oneway) %>%
  mutate(resolve_speed_limit(highway, maxspeed_raw))

st_write(study_area_streets, "data/infrastructure/study_area_streets.gpkg", delete_dsn = TRUE)
```


## Annual Average Daily Traffic (IDOT)

Annual average daily traffic (AADT) estimates come from the Illinois Department of Transportation's (IDOT) [Annual Average Daily Traffic layer](https://gis-idot.opendata.arcgis.com/datasets/annual-average-daily-traffic-aadt), downloaded as GeoJSON to `data/infrastructure/aadt.geojson`. AADT is the estimated number of vehicles that travel a road segment on an average day of the year, in both directions. IDOT produces these estimates through its Traffic Count Program, which combines continuous counts from permanent stations with short-term counts, usually 24 hours long, taken at thousands of locations on a rotating multi-year schedule and adjusted to an annual average. Each feature is a road segment that records total daily traffic (`AADT`) and the year of that estimate (`AADT_YR`), along with daily heavy commercial vehicle volumes (`HCV_AADT`), split into single-unit trucks and buses (`SU_AADT`) and multi-unit combination trucks (`MU_AADT`), and the year of the truck estimate (`HCV_AADTYR`). The remaining columns identify the segment: IDOT's route inventory key (`INVENTORY`) and key route number (`KEY_RT_NBR`), the signed route name (`MARKED_NAM`), the local street name (`ROAD_NAME`), the county (`COUNTY_NAM`), and the segment's original length (`Shape_Length`).

A few points matter when using these data. Count years differ by segment, so a single download mixes estimates from several years. Coverage is strongest on state and major roads, and many local streets have no value, which should be treated as missing rather than zero. `Shape_Length` reflects the unclipped segment, so length is recalculated with `st_length()` after clipping to Cook County. In this project, AADT serves as a measure of the vehicle traffic pedestrians are exposed to at intersections and crosswalks. The column definitions above are informal and should be checked against the layer's metadata on the IDOT portal.

```{r}
aadt <- st_read("data/infrastructure/aadt.geojson") %>% 
  st_transform(3435) %>%
  st_make_valid() %>%
  st_intersection(st_geometry(cc_boundary)) %>% 
  st_as_sf() %>% 
  mutate(miles = st_length(geometry),
         miles = set_units(miles, mi))

st_write(aadt, "data/infrastructure/study_area_aadt.gpkg", delete_dsn = TRUE)
```

## Pedestrian Fatalities (FARS)

Fatal crash records for Cook County from 2010 to 2024 are pulled from NHTSA's Fatality Analysis Reporting System (FARS). The records are narrowed to crashes whose first harmful event involved a pedestrian. Crashes with missing or placeholder coordinates are dropped so that each remaining fatality can be mapped at its reported location. A readable date and time is added to each record for display on the map, using whichever date and time fields the data provides.

```{r}
#| label: crashes-fars

# Cook County, Illinois FIPS codes -- fixed constants for this project's
# single-county scope.
state_fips  <- 17L  # Illinois
county_fips <- 31L  # Cook County

crashes_fars <- read_csv(paste0(
  "https://crashviewer.nhtsa.dot.gov/crashviewer/CrashAPI/FARSData/GetFARSData?",
  "dataset=Accident&FromYear=2010&ToYear=2024&State=", state_fips,
  "&county=", county_fips, "&format=csv"
), show_col_types = FALSE) %>%
  filter(state == state_fips, county == county_fips)

ped_fatalities <- crashes_fars %>%
  filter(harm_evname == "Pedestrian") %>%
  filter(
    latitude  > -90,  latitude  < 90,
    longitud  > -180, longitud  < 180,
    !latitude  %in% c(77.7777, 88.8888, 99.9999),   # FARS sentinels for unknown/not-reported coordinates
    !longitud  %in% c(-77.7777, -88.8888, -99.9999)
  ) %>%
  st_as_sf(coords = c("longitud", "latitude"), crs = 4326)

# Build hover date/time text defensively -- check what's actually present
# in the CrashAPI response rather than assuming specific column names.
get_col_or_null <- function(df, col) if (col %in% names(df)) df[[col]] else NULL

monthname <- get_col_or_null(ped_fatalities, "monthname")
month_num <- get_col_or_null(ped_fatalities, "month")
day       <- get_col_or_null(ped_fatalities, "day")
year      <- get_col_or_null(ped_fatalities, "year")
hourname  <- get_col_or_null(ped_fatalities, "hourname")
hour_num  <- get_col_or_null(ped_fatalities, "hour")
minute    <- get_col_or_null(ped_fatalities, "minute")

fatality_date <- if (!is.null(monthname) && !is.null(day) && !is.null(year)) {
  paste(monthname, day, year)
} else if (!is.null(month_num) && !is.null(day) && !is.null(year)) {
  sprintf("%02d/%02d/%d", month_num, day, year)
} else {
  NA_character_
}

fatality_time <- if (!is.null(hourname)) {
  as.character(hourname)
} else if (!is.null(hour_num) && !is.null(minute)) {
  sprintf("%02d:%02d", hour_num, minute)
} else {
  NA_character_
}

ped_fatalities_formatted <- ped_fatalities %>%
  # Attach fatality_date/fatality_time as real columns before the join below,
  # rather than leaving them as loose chunk-local vectors -- otherwise
  # they never make it into the saved .gpkg (and, since drop_na(geoid) below
  # removes rows, couldn't safely be attached after the join anyway).
  filter(caseyear >= 2022) %>% 
  st_join(study_area) %>%
  drop_na(geoid) %>%
  st_make_valid() %>%
  st_as_sf()

st_write(ped_fatalities_formatted, "data/crashes/ped_fatalities.gpkg", delete_dsn = TRUE)
```

## Crash Records (IDOT)

Crash records for 2022 to 2024 are read from IDOT's annual crash files and combined into a single dataset. Each crash is mapped using its recorded coordinates, falling back to a second set of coordinates when the first is missing or falls outside Illinois. The date and time are built from the separate year, month, day and hour fields, since the combined date field can shift late-evening crashes to the next day. Each crash is flagged as a pedestrian crash where applicable, and as a killed-or-seriously-injured (KSI) crash if anyone died or was seriously injured; a second output contains only the pedestrian crashes.

```{r}
#| label: idot-crashes

# Required lowercased column names this function depends on -- checked
# explicitly below so a schema change in a future IDOT export (a renamed
# or dropped column) fails with a clear message naming the column, rather
# than a cryptic downstream "object not found" error from dplyr.
required_idot_cols <- c(
  "x", "y", "tscrashlongitude", "tscrashlatitude",
  "crashyr", "crashmonth", "crashday", "crashhour",
  "typeoffirstcrash", "totalfatals","totalinjured", "ainjuries"
)

read_crash_year <- function(year, dir) {
  file <- list.files(dir, pattern = paste0(year, ".*\\.csv$"),
                     full.names = TRUE, ignore.case = TRUE)
  if (length(file) == 0) {
    warning("No CSV found for ", year, " in ", dir, "; skipping.")
    return(NULL)
  }
  if (length(file) > 1) {
    warning("Multiple CSVs match ", year, "; using ", basename(file[1]))
    file <- file[1]
  }
  read_csv(file, col_types = cols(.default = col_character()),
           show_col_types = FALSE) %>%
    rename_with(tolower) %>%
    mutate(source_file = basename(file))
}

process_crashes <- function(years, dir, crs_out = 4326, ped_only = FALSE) {
  crash_raw <- years %>%
    map(read_crash_year, dir = dir) %>%
    compact() %>%
    bind_rows() %>%
    type_convert(col_types = cols())

  missing_cols <- setdiff(required_idot_cols, names(crash_raw))
  if (length(missing_cols) > 0) {
    stop(
      "IDOT crash CSV is missing expected column(s): ",
      paste(missing_cols, collapse = ", "),
      ". Check the raw column headers (rename_with(tolower) is applied) ",
      "and update required_idot_cols/process_crashes() to match."
    )
  }

  in_il <- function(lon, lat) {
    !is.na(lon) & !is.na(lat) &
      between(lon, -91.6, -87.0) & between(lat, 36.9, 42.6)
  }

  # x/y are the ArcGIS-exported coordinate pair, distinct from
  # tscrashcoordinatex/y (a separate, likely projected pair not used here)
  # and from tscrashlongitude/tscrashlatitude (the fallback). Report how
  # often x/y actually resolves as in-Illinois lon/lat vs. falls back, as a
  # sanity check that x/y really is geographic and not itself projected.
  n_xy_valid <- sum(in_il(crash_raw$x, crash_raw$y), na.rm = TRUE)
  cli::cli_alert_info(
    "IDOT crashes: {n_xy_valid} of {nrow(crash_raw)} rows used x/y directly (rest fell back to TSCrashLongitude/TSCrashLatitude)."
  )

  crash_df <- crash_raw %>%
    mutate(
      lon = if_else(in_il(x, y), x, tscrashlongitude),
      lat = if_else(in_il(x, y), y, tscrashlatitude),
      # crash date/time built from crashyr/crashmonth/crashday/crashhour
      # rather than the raw crashdate field, which is UTC-shifted and can
      # move late-evening crashes to the next calendar day.
      crash_date = make_date(crashyr + 2000, crashmonth, crashday),
      crash_dth   = crash_date + hours(crashhour),
      ped = case_when(
        typeoffirstcrash == "Pedestrian"   ~ "Pedestrian",
        TRUE                               ~ NA_character_
      ),
      ksi = totalfatals > 0 | ainjuries > 0
    ) %>%
    select(-any_of("crash_hour"))

  crash <- crash_df %>%
    filter(in_il(lon, lat)) %>%
    st_as_sf(coords = c("lon", "lat"), crs = 4326, remove = FALSE)

  if (ped_only) crash <- filter(crash, !is.na(ped))

  st_transform(crash, crs_out)
}

idot_years <- 2022:2024
idot_dir   <- "data/crashes"

il_crashes <- process_crashes(idot_years, dir = idot_dir)
il_ped_crashes <- process_crashes(idot_years, dir = idot_dir, ped_only = TRUE)

# join crashes with study area
cc_crashes <- il_crashes %>%
  select(
    x = lon,
    y = lat,
    year = crashyr,
    month = crashmonth,
    day = crashday,
    datetime = crash_dth,
    crash_type = collisiontypecode,
    crash_first = typeoffirstcrash,
    fatals = totalfatals,
    injurs = totalinjured,
    ainjurs = ainjuries,
    binjurs = binjuries,
    cinjurs = cinjuries,
    ksi,
    severity = crashseverity,
  ) %>%
  st_transform(crs = 3435) %>%
  st_join(study_area) %>%
  drop_na(geoid) %>%
  st_make_valid() %>%
  st_as_sf() 

# join ped crashes with study area
cc_ped_crashes <- il_ped_crashes %>%
  select(
    x = lon,
    y = lat,
    year = crashyr,
    month = crashmonth,
    day = crashday,
    datetime = crash_dth,
    crash_type = collisiontypecode,
    ped,
    fatals = totalfatals,
    injurs = totalinjured,
    ainjurs = ainjuries,
    binjurs = binjuries,
    cinjurs = cinjuries,
    ksi,
    severity = crashseverity,
  ) %>% 
  st_transform(crs = 3435) %>% 
  st_join(study_area) %>% 
  drop_na(geoid) %>% 
  st_make_valid() %>% 
  st_as_sf() 

# write to folder
st_write(obj = cc_crashes, dsn = "data/crashes/cc_crashes_2022_2024.gpkg", append = FALSE)
st_write(cc_ped_crashes, "data/crashes/cc_ped_crashes_2022_2024.gpkg", append = FALSE)
```

## Area Deprivation Index (ADI)

This section produces the neighborhood disadvantage measures used in the Socioeconomic and Contextual Data table for Aim 2. It calculates the Area Deprivation Index (ADI), a composite score of neighborhood socioeconomic disadvantage based on income, education, employment, and housing, for two sets of areas: Chicago's community areas, built up from 2020 census tracts, and the municipalities of suburban Cook County. The underlying data come from the 2024 American Community Survey 5-year estimates, and the approach is adapted from an earlier analysis in adi_calculations.qmd.

The sociome R package usually handles ADI calculations automatically, but its standard function only works with built-in geographies such as states, counties, and census tracts. It can't produce scores for municipalities or for custom areas like Chicago community areas. To work around this, the code downloads the census data separately and passes it to the package's core ADI calculation step, a route the package's documentation supports. It requests the same census variables the package would use on its own, so the results follow the same method as standard ADI scores.

Unlike the national 1–100 percentiles published by the Neighborhood Atlas, these scores are relative to the areas included in the calculation. Higher values indicate greater disadvantage.

```{r}
#| label: adi-setup
# Set once per machine: tidycensus::census_api_key(Sys.getenv("CENSUS_API_KEY"), install = TRUE)
CENSUS_API_KEY <- Sys.getenv("CENSUS_API_KEY", unset = "8f6a0a83c8a2466e3e018a966846c86412d0bb6e")
tidycensus::census_api_key(CENSUS_API_KEY, install = TRUE, overwrite = TRUE)

adi_year   <- 2024
adi_survey <- "acs5"

adi_acs_vars <- c(
  "B11005_001", "B15003_001", "B15003_002", "B15003_003", "B15003_004",
  "B15003_005", "B15003_006", "B15003_007", "B15003_008", "B15003_009",
  "B15003_010", "B15003_011", "B15003_012", "B15003_017", "B15003_018",
  "B15003_019", "B15003_020", "B15003_021", "B15003_022", "B15003_023",
  "B15003_024", "B15003_025", "B17010_001", "B17010_002", "B19001_002",
  "B19001_011", "B19001_012", "B19001_013", "B19001_014", "B19001_015",
  "B19001_016", "B19001_017", "B19113_001", "B11005_002", "B11005_005",
  "B23025_003", "B23025_005", "B25003_001", "B25003_002", "B25014_001",
  "B25014_005", "B25014_006", "B25014_007", "B25014_011", "B25014_012",
  "B25014_013", "B25044_001", "B25044_003", "B25044_010", "B25064_001",
  "B25077_001", "B25088_002", "C17002_001", "C17002_002", "C17002_003",
  "C17002_004", "C17002_005", "C24010_001", "C24010_003", "C24010_039"
)

adi_median_vars <- c("B19113_001", "B25064_001", "B25077_001", "B25088_002")
adi_sum_vars    <- setdiff(adi_acs_vars, adi_median_vars)

get_acs_tidy_wide <- function(..., variables) {
  tidycensus::get_acs(
    ...,
    variables = variables,
    year = adi_year,
    survey = adi_survey,
    output = "tidy"
  ) %>%
    select(GEOID, NAME, variable, estimate) %>%
    pivot_wider(names_from = variable, values_from = estimate)
}
```

### Chicago community area

This step calculates ADI for each of Chicago's 77 community areas. Census data aren't published by community area, so the code starts with 2024 American Community Survey estimates for every census tract in Cook County, using 2020 tract boundaries. It assigns each tract to the community area it falls within, using the tract boundary file from the City of Chicago Data Portal and the community area boundaries defined earlier in the Study Area Geography section. It then combines the tract data within each community area and calculates one ADI score per area.

```{r}
#| label: adi-community-areas

tracts_raw <- get_acs_tidy_wide(
  geography = "tract",
  state = "17",
  county = "031",
  variables = c(adi_acs_vars, "B01003_001")  # + total population, to weight the medians below
)

tract_boundaries <- st_read(
  "https://data.cityofchicago.org/resource/7qjk-677h.geojson",
  quiet = TRUE
) %>%
  st_transform(3435)

tract_geoid_col <- intersect(
  names(tract_boundaries), c("geoid20", "GEOID20", "geoid", "GEOID")
)[1]

tract_ca_lookup <- tract_boundaries %>%
  mutate(GEOID = tract_full) %>%
  st_point_on_surface() %>%
  st_join(community_areas %>% select(commarea_geoid = geoid, commarea_name = name)) %>%
  st_drop_geometry() %>%
  select(GEOID, commarea_geoid, commarea_name) %>%
  filter(!is.na(commarea_geoid))  # tract centroids outside any community area

tracts_with_ca <- tracts_raw %>%
  mutate(GEOID = as.character(GEOID)) %>%
  inner_join(tract_ca_lookup, by = "GEOID")

adi_ca_agg <- tracts_with_ca %>%
  group_by(commarea_geoid, commarea_name) %>%
  summarise(
    across(all_of(adi_sum_vars), \(x) sum(x, na.rm = TRUE)),
    across(all_of(adi_median_vars), \(x) weighted.mean(x, w = B01003_001, na.rm = TRUE)),
    .groups = "drop"
  ) %>%
  rename(GEOID = commarea_geoid, NAME = commarea_name) %>%
  relocate(GEOID, NAME)

adi_community_areas <- calculate_adi(adi_ca_agg, keep_indicators = TRUE)

write_csv(adi_community_areas, "data/census/adi_community_areas.csv")
```

### Suburban municipality

This step calculates ADI for each municipality in suburban Cook County. The Census Bureau publishes data for cities, villages, and towns as "places," but it can only provide them for an entire state at once, not by county. The code therefore downloads 2024 American Community Survey data for all of Illinois's roughly 1,300 places and keeps only the Cook County municipalities identified in the Study Area Geography section. Filtering before the calculation matters because ADI scores are relative to the set of areas they are computed from. Each suburb is therefore scored against other suburban Cook County municipalities, not against communities statewide.

```{r}
#| label: adi-places

places_raw <- get_acs_tidy_wide(
  geography = "place",
  state = "17",
  variables = adi_acs_vars
)

suburban_place_geoids <- places_cook_county$geoid

places_cc_raw <- places_raw %>%
  mutate(GEOID = as.character(GEOID)) %>%
  filter(GEOID %in% suburban_place_geoids)

# places_cook_county$geoid's exact provenance (Metopio's place topojson,
# Study Area Geography above) wasn't independently verified as matching
# tidycensus's Census Place GEOIDs from this environment -- if this warns,
# check that assumption before trusting adi_places.
if (nrow(places_cc_raw) < length(suburban_place_geoids) * 0.9) {
  cli::cli_alert_warning(
    "Matched only {nrow(places_cc_raw)} of {length(suburban_place_geoids)} suburban municipalities to Census place GEOIDs."
  )
}

adi_places <- calculate_adi(places_cc_raw, keep_indicators = TRUE)

write_csv(adi_places, "data/census/adi_suburban_places.csv")
```

### Study Area ADI
```{r}
adi_study_area <- adi_community_areas %>% 
  bind_rows(adi_places)

write_csv(adi_study_area, "data/census/adi_study_area.csv")
```

## Read, Format and Save

### Read datasets
```{r}

county_boundary <- st_read(dsn = "data/boundaries/cc_boundary.gpkg") %>% st_transform(crs=4326)
cog_boundaries <- st_read(dsn = "data/boundaries/cogs_boundaries.gpkg") %>% st_transform(crs=4326)
study_area <- st_read(dsn = "data/boundaries/study_area.gpkg") %>% st_transform(crs=4326)
intersections <- st_read("data/intersections/study_area_intersections.gpkg") %>% st_transform(crs=4326)
streets <- st_read("data/infrastructure/study_area_streets.gpkg") %>% st_transform(crs=4326)
walkability <- st_read("data/infrastructure/walkability.gpkg") %>% st_transform(crs=4326)
ped_fatalities <- st_read("data/crashes/ped_fatalities.gpkg") %>% st_transform(crs=4326)
ped_crashes <- st_read("data/crashes/cc_ped_crashes_2022_2024.gpkg") %>% st_transform(crs=4326)
adi <- read.csv("data/census/adi_study_area.csv")
aadt <- st_read("data/infrastructure/study_area_aadt.gpkg") %>% st_transform(crs=4326)
land_use <- st_read("data/boundaries/land_use_simplified_2020.gpkg") %>% st_transform(crs=4326)

```

Save the assembled `sf`/`tibble` tables to a single `.RData` file. [Dataset Development](phase01.html) loads this file to build an interactive data-exploration map; downstream analysis in [Data Presentation and Analysis](phase02.html) will read from these same tables (or their saved shapefiles/CSVs under `output/`).

```{r}
#| label: save-dataprep

save(
  study_area,
  county_boundary,
  cog_boundaries,
  streets,
  intersections,
  walkability,
  ped_fatalities,
  ped_crashes,
  adi,
  aadt,
  # land_use,
  file = "data/phase01_dataprep.RData"
)
```

