---
title: "Data Preparation for Phase 1 — Dataset Development"
subtitle: "Street Design as a Public Health Determinant: Crosswalk Geometry and Pedestrian Injury Risk Across Cook County"
date: today
format:
  html:
    toc: true
    toc-depth: 2
    code-fold: true
    number-sections: false
    theme: cosmo
    embed-resources: true
    warning: false
editor: source
execute:
  eval: false
editor_options:
  chunk_output_type: console
---

## Purpose

This document builds the `sf` data tables that support [`phase01.qmd`](phase01.qmd)'s dataset-development tasks: the countywide study-area geography, street intersections, the street/speed-limit network, pedestrian fatality records (NHTSA FARS), and crash records (IDOT). It ends by saving the resulting tables to `data/phase1_dataprep.RData`, which `phase01.qmd` loads (a subset of) to build an interactive map for data exploration.

This is dataset-creation code, not a narrative workflow — for first-time environment setup (conda/osmnx), see [`Phase0.qmd`](Phase0.qmd)'s Prerequisites section; this document assumes that `osmnx-env` conda environment already exists.

## Setup

```{r}
#| label: r-packages
#| eval: true

library(sf)
library(reticulate)
library(dplyr)
library(tidyr)
library(purrr)
library(readr)
library(glue)
library(cli)
library(tigris)
library(units)
library(osmextract)
library(lubridate)
library(stringr)
```

```{r}
#| label: osmnx-setup

reticulate::use_condaenv("osmnx-env", required = TRUE)

# convert = FALSE: keep everything returned by these modules as live Python
# object references rather than letting reticulate auto-convert them to R
# the instant a call returns (see get_area_intersections() below for why).
ox <- reticulate::import("osmnx", convert = FALSE)
shapely_wkt <- reticulate::import("shapely.wkt", convert = FALSE)

ox$settings$use_cache <- TRUE
ox$settings$log_console <- TRUE
```

## Study Area Geography

Supports `phase01.qmd`'s *Study Area Geography* table. Combines the City of Chicago's 77 community areas with suburban Cook County's incorporated municipalities (≥4 sq mi) into a single, internally consistent geography (`study_area`, N ≈ 200), identified by `geoid` and `name`, with a `type` field distinguishing the two.

```{r}
#| label: study-area

counties_il <- counties(state = "17", cb = TRUE)

cc_boundary <- counties_il %>%
  filter(GEOID == "17031") %>%
  st_transform(3435) %>%
  select(geoid_county = GEOID, name_county = NAME,
         area_land = ALAND, area_water = AWATER) %>%
  mutate(
    area_land_sqmi  = set_units(set_units(area_land, m^2), mi^2),
    area_water_sqmi = set_units(set_units(area_water, m^2), mi^2),
    area_sqmi = area_land_sqmi + area_water_sqmi
  ) %>%
  select(-area_land, -area_water)

st_write(cc_boundary, "data/boundaries/cc_boundary.shp", delete_dsn = TRUE)

community_areas <- st_read(
  "https://metopio.blob.core.windows.net/lalage/atlas/3/shapes/communityareas.topo.json",
  quiet = TRUE
) %>%
  st_as_sf() %>%
  st_set_crs(4326) %>%
  st_transform(3435) %>%
  rename(geoid = id) %>%
  mutate(
    geoid = str_replace(geoid, "1714000-", ""),
    location = "Chicago, IL",
    type = "community area"
  )

places_cook_county <- st_read(
  "https://metopio.blob.core.windows.net/lalage/atlas/6/shapes/SCC-place.topo.json",
  quiet = TRUE
) %>%
  st_as_sf() %>%
  st_set_crs(4326) %>%
  st_transform(3435) %>%
  rename(geoid = id) %>%
  mutate(
    sqmi = as.numeric(st_area(geometry)) * 0.0000003861,
    location = "suburban Cook County",
    type = "municipality"
  ) %>%
  filter(sqmi >= 4) %>%
  select(-sqmi)

study_area <- places_cook_county %>%
  bind_rows(community_areas) %>%
  st_as_sf() %>%
  st_make_valid()

st_write(study_area, "data/boundaries/study_area.shp", delete_dsn = TRUE)

# Re-project back to unprojected lat/lon: osmnx (Step "Street Intersections",
# below) and leaflet (phase01.qmd's map) both expect EPSG:4326, not the
# EPSG:3435 (feet) projection used above for area/distance calculations.
cc_boundary <- st_transform(cc_boundary, 4326)
study_area  <- st_transform(study_area, 4326)
```

::: {.callout-note collapse="true"}
## Note: superseding the pre-community-area run
An earlier version of this pipeline built the run list from `cc_municipalities.shp` (suburban municipalities plus a single, undivided "Chicago" polygon, columns `muni_id`/`muni_name`) rather than from `study_area` above. The already-committed `output/all_intersections.shp` and `config/cc_muni_run_list.csv` reflect that earlier geography. The run list, intersection function, and downstream tables below use `study_area`'s `geoid`/`name` columns instead, and its **77 separate Chicago community areas** rather than one citywide polygon — re-running Step 3 onward is needed to bring the committed intersection output in line with the current study-area definition.
:::

## Street Intersections

Supports `phase01.qmd`'s *Street Intersections* table (candidate crossing locations for Aim 1; spatial units for Aims 3–4).

Road-class filter shared with the Streets section below, matching the reference pipeline's `osm_ingest.py` (lines 51–71, 127):

```{r}
#| label: filter-constants

highway_subtypes <- c(
  "primary", "primary_link", "unclassified", "residential",
  "secondary", "secondary_link", "service", "tertiary",
  "tertiary_link", "trunk", "trunk_link", "busway",
  "construction", "corridor"
)
service_exclusion <- c("parking_aisle", "driveway")

drive_filter <- glue(
  '["highway"~"{paste(highway_subtypes, collapse = "|")}"]',
  '["service"!~"{paste(service_exclusion, collapse = "|")}"]',
  '["access"!~"private"]["aeroway"!~".*"]["indoor"!~"yes"]'
)
```

Build the per-area run list — one row per `study_area` unit, with a WKT polygon and a `wave` number for batched, resumable processing:

```{r}
#| label: build-run-list

N_PER_WAVE <- 12  # tune based on how much Overpass load you want per batch

run_list <- study_area %>%
  st_transform(3435) %>%                       # IL State Plane East (ft), planar repair
  st_make_valid() %>%
  st_collection_extract("POLYGON") %>%         # drop stray lines/points from the repair
  group_by(geoid, name, type) %>%
  summarise(geometry = st_union(geometry), .groups = "drop") %>%
  st_transform(4326) %>%                       # OSMnx expects lat/lon
  mutate(wkt = st_as_text(geometry, digits = 15)) %>%
  st_drop_geometry() %>%
  arrange(name) %>%
  mutate(wave = ceiling(row_number() / N_PER_WAVE))

dir.create("config", showWarnings = FALSE)
write_csv(run_list, "config/cc_run_list.csv")
```

Extract OSMnx-consolidated drive-network intersections for one study-area unit:

```{r}
#| label: intersection-function

get_area_intersections <- function(geoid, name, wkt,
                                    retain_all = FALSE, tolerance = NULL) {
  poly <- shapely_wkt$loads(wkt)

  G_drive <- tryCatch(
    ox$graph_from_polygon(
      poly,
      network_type = "drive",
      custom_filter = drive_filter,
      retain_all = retain_all
    ),
    error = function(e) {
      cli::cli_alert_warning("{name}: {conditionMessage(e)}")
      NULL
    }
  )
  if (is.null(G_drive)) return(NULL)

  G_drive <- ox$project_graph(G_drive)

  consolidate_args <- list(G = G_drive, rebuild_graph = FALSE, dead_ends = FALSE)
  if (!is.null(tolerance)) consolidate_args$tolerance <- tolerance
  intersections <- do.call(ox$consolidate_intersections, consolidate_args)

  # intersections is a non-converting (convert = FALSE) Python GeoSeries --
  # extract x/y as numpy arrays (.to_numpy()) rather than converting the
  # pandas object itself, which is the operation that crashes reticulate.
  coords_df <- intersections$to_crs("EPSG:4326")$get_coordinates()
  lon <- as.numeric(reticulate::py_to_r(coords_df$x$to_numpy()))
  lat <- as.numeric(reticulate::py_to_r(coords_df$y$to_numpy()))

  tibble(lon = lon, lat = lat, geoid = geoid, name = name)
}
```

Run across all study-area units, wave by wave. This queries the shared public Overpass API underneath `osmnx` — process sequentially within each wave, with a short pause between units, and skip units already written so a failed run can resume without redoing completed ones. See `Phase0.qmd`'s Step 6 for a list of alternate Overpass mirrors if the default endpoint is timing out.

```{r}
#| label: run-waves

out_dir <- "data/intersections/raw"
ox$settings$overpass_url <- "https://overpass-api.de/api/interpreter"
dir.create(out_dir, recursive = TRUE, showWarnings = FALSE)

run_wave <- function(wave_df, pause_sec = 1) {
  purrr::pwalk(wave_df, function(geoid, name, wkt, wave, ...) {
    out_file <- file.path(out_dir, glue("{geoid}.csv"))
    if (file.exists(out_file)) return(invisible(NULL))  # resume support

    cli::cli_progress_step("Processing {name}")
    result <- get_area_intersections(geoid, name, wkt)
    if (!is.null(result)) readr::write_csv(result, out_file)
    Sys.sleep(pause_sec)
  })
}

run_list %>%
  group_split(wave) %>%
  walk(run_wave)
```

Combine per-unit results into a single countywide layer:

```{r}
#| label: combine-intersections

intersection_files <- list.files(out_dir, full.names = TRUE)
all_intersections <- map_dfr(intersection_files, read_csv, show_col_types = FALSE)

all_intersections_sf <- st_as_sf(all_intersections, coords = c("lon", "lat"), crs = 4326)
st_write(all_intersections_sf, "output/all_intersections.shp", delete_dsn = TRUE)
```

## Streets and Speed Limits

Supports `phase01.qmd`'s *Streets and Speed Limits* table (roadway functional class/speed limit as control variables for Aims 3–4).

**Speed-limit resolution:** OSM's `maxspeed` tag carries a real posted speed limit where mapped, but coverage is uneven. Where it's missing, fall back to a default by OSM `highway` class. These class defaults are a heuristic, not measured data — Illinois' statutory default for an urban district absent posted signage is 30 mph (625 ILCS 5/11-601(b)); many Cook County suburbs post/ordinance residential streets down to 25 mph, used here as the more common default for `residential`/`living_street`. `maxspeed_source` records which applied per segment.

```{r}
#| label: speed-limit-defaults

default_speed_by_highway <- c(
  motorway = 65, motorway_link = 45,
  trunk = 55, trunk_link = 40,
  primary = 45, primary_link = 35,
  secondary = 35, secondary_link = 30,
  tertiary = 30, tertiary_link = 25,
  unclassified = 30,
  residential = 25,
  living_street = 15,
  service = 15
)

# OSM tag values can themselves be semicolon-joined (e.g. "30;20") -- use
# the first for class-default lookup and mph parsing.
first_tag_value <- function(x) sub(";.*$", "", x)

parse_maxspeed_mph <- function(x) {
  first <- first_tag_value(x)
  # substr()/regexpr() only, not regmatches() -- regmatches() with a single
  # regexpr() result silently drops non-matching elements rather than
  # returning a same-length vector, which would misalign this against the
  # rest of the tibble whenever maxspeed is untagged (the common case).
  m <- regexpr("[0-9]+", first)
  starts <- as.integer(m)
  lens <- attr(m, "match.length")
  digits <- ifelse(starts > 0, substr(first, starts, starts + lens - 1L), NA_character_)
  suppressWarnings(as.numeric(digits))
}

resolve_speed_limit <- function(highway, maxspeed_raw) {
  tagged_mph <- parse_maxspeed_mph(maxspeed_raw)
  class_default <- unname(default_speed_by_highway[first_tag_value(highway)])
  tibble(
    maxspeed_mph = ifelse(!is.na(tagged_mph), tagged_mph, class_default),
    maxspeed_source = ifelse(!is.na(tagged_mph), "osm_tagged", "class_default")
  )
}
```

**Streets, from OSM via Geofabrik:** a bulk offline extract, filtered to road ways during the GDAL `.pbf` → `.gpkg` conversion itself (much faster than reading all of Illinois' line features and filtering afterward), then clipped to the county boundary.

```{r}
#| label: streets-osm
#| # https://download.geofabrik.de/north-america/us/illinois.html

il_streets <- oe_read(
  "https://download.geofabrik.de/north-america/us/illinois-latest.osm.pbf",
  layer = "lines",
  extra_tags = c("maxspeed", "oneway"),
  query = "SELECT * FROM lines WHERE highway IS NOT NULL",
  quiet = TRUE
)

cc_streets <- il_streets %>%
  filter(highway %in% highway_subtypes) %>%
  st_transform(4326) %>%
  st_make_valid() %>%
  st_intersection(st_geometry(cc_boundary)) %>%
  st_collection_extract("LINESTRING") %>%
  select(street_name = name, highway, maxspeed_raw = maxspeed, oneway) %>%
  mutate(resolve_speed_limit(highway, maxspeed_raw))

st_write(cc_streets, "output/all_streets.shp", delete_dsn = TRUE)
```

## Pedestrian Fatalities (FARS)

Supports `phase01.qmd`'s *Pedestrian Fatalities (FARS)* table (Aim 3).

```{r}
#| label: crashes-fars

# Cook County, Illinois FIPS codes -- fixed constants for this project's
# single-county scope.
state_fips  <- 17L  # Illinois
county_fips <- 31L  # Cook County

crashes_fars <- read_csv(paste0(
  "https://crashviewer.nhtsa.dot.gov/crashviewer/CrashAPI/FARSData/GetFARSData?",
  "dataset=Accident&FromYear=2010&ToYear=2024&State=", state_fips,
  "&county=", county_fips, "&format=csv"
), show_col_types = FALSE) %>%
  filter(state == state_fips, county == county_fips)

ped_fatalities <- crashes_fars %>%
  filter(harm_evname == "Pedestrian") %>%
  filter(
    latitude  > -90,  latitude  < 90,
    longitud  > -180, longitud  < 180,
    !latitude  %in% c(77.7777, 88.8888, 99.9999),   # FARS sentinels for unknown/not-reported coordinates
    !longitud  %in% c(-77.7777, -88.8888, -99.9999)
  ) %>%
  st_as_sf(coords = c("longitud", "latitude"), crs = 4326)

# Build hover date/time text defensively -- check what's actually present
# in the CrashAPI response rather than assuming specific column names.
get_col_or_null <- function(df, col) if (col %in% names(df)) df[[col]] else NULL

monthname <- get_col_or_null(ped_fatalities, "monthname")
month_num <- get_col_or_null(ped_fatalities, "month")
day       <- get_col_or_null(ped_fatalities, "day")
year      <- get_col_or_null(ped_fatalities, "year")
hourname  <- get_col_or_null(ped_fatalities, "hourname")
hour_num  <- get_col_or_null(ped_fatalities, "hour")
minute    <- get_col_or_null(ped_fatalities, "minute")

fatality_date <- if (!is.null(monthname) && !is.null(day) && !is.null(year)) {
  paste(monthname, day, year)
} else if (!is.null(month_num) && !is.null(day) && !is.null(year)) {
  sprintf("%02d/%02d/%d", month_num, day, year)
} else {
  NA_character_
}

fatality_time <- if (!is.null(hourname)) {
  as.character(hourname)
} else if (!is.null(hour_num) && !is.null(minute)) {
  sprintf("%02d:%02d", hour_num, minute)
} else {
  NA_character_
}

ped_fatalities <- ped_fatalities %>%
  mutate(fatality_date = fatality_date, fatality_time = fatality_time)

st_write(ped_fatalities, "output/ped_fatalities.shp", delete_dsn = TRUE)
```

## Crash Records (IDOT)

Supports `phase01.qmd`'s *Crash Records (IDOT)* table (Aim 4). Stacks yearly IDOT crash-report CSV exports, cleans coordinates and dates, and identifies pedestrian/bicycle crashes and KSI (killed-or-seriously-injured) severity. Adapted from [`scripts/QMD/idot_crashes.qmd`](scripts/QMD/idot_crashes.qmd), pointed at this project's `data/boundaries/CRASHES_*.csv` files rather than a parameterized directory.

Confirmed against the actual exported column headers (`CRASHES_2022/2023/2024.csv`; PascalCase, no underscores except `City_Township_Flag`/`Crash_Hour`) — after `rename_with(tolower)`, every column this code reads (`x`, `y`, `tscrashlongitude`, `tscrashlatitude`, `crashyr`, `crashmonth`, `crashday`, `crashhour`, `typeoffirstcrash`, `totalfatals`, `ainjuries`, `crashseverity`, `crash_hour`) is present as-is; no renaming was needed. `crash_hour` (a near-duplicate of `crashhour`) is dropped rather than used.

```{r}
#| label: idot-crashes

# Required lowercased column names this function depends on -- checked
# explicitly below so a schema change in a future IDOT export (a renamed
# or dropped column) fails with a clear message naming the column, rather
# than a cryptic downstream "object not found" error from dplyr.
required_idot_cols <- c(
  "x", "y", "tscrashlongitude", "tscrashlatitude",
  "crashyr", "crashmonth", "crashday", "crashhour",
  "typeoffirstcrash", "totalfatals", "ainjuries"
)

read_crash_year <- function(year, dir) {
  file <- list.files(dir, pattern = paste0(year, ".*\\.csv$"),
                     full.names = TRUE, ignore.case = TRUE)
  if (length(file) == 0) {
    warning("No CSV found for ", year, " in ", dir, "; skipping.")
    return(NULL)
  }
  if (length(file) > 1) {
    warning("Multiple CSVs match ", year, "; using ", basename(file[1]))
    file <- file[1]
  }
  read_csv(file, col_types = cols(.default = col_character()),
           show_col_types = FALSE) %>%
    rename_with(tolower) %>%
    mutate(source_file = basename(file))
}

process_crashes <- function(years, dir, crs_out = 4326, ped_bike_only = FALSE) {
  crash_raw <- years %>%
    map(read_crash_year, dir = dir) %>%
    compact() %>%
    bind_rows() %>%
    type_convert(col_types = cols())

  missing_cols <- setdiff(required_idot_cols, names(crash_raw))
  if (length(missing_cols) > 0) {
    stop(
      "IDOT crash CSV is missing expected column(s): ",
      paste(missing_cols, collapse = ", "),
      ". Check the raw column headers (rename_with(tolower) is applied) ",
      "and update required_idot_cols/process_crashes() to match."
    )
  }

  in_il <- function(lon, lat) {
    !is.na(lon) & !is.na(lat) &
      between(lon, -91.6, -87.0) & between(lat, 36.9, 42.6)
  }

  # x/y are the ArcGIS-exported coordinate pair, distinct from
  # tscrashcoordinatex/y (a separate, likely projected pair not used here)
  # and from tscrashlongitude/tscrashlatitude (the fallback). Report how
  # often x/y actually resolves as in-Illinois lon/lat vs. falls back, as a
  # sanity check that x/y really is geographic and not itself projected.
  n_xy_valid <- sum(in_il(crash_raw$x, crash_raw$y), na.rm = TRUE)
  cli::cli_alert_info(
    "IDOT crashes: {n_xy_valid} of {nrow(crash_raw)} rows used x/y directly (rest fell back to TSCrashLongitude/TSCrashLatitude)."
  )

  crash_df <- crash_raw %>%
    mutate(
      lon = if_else(in_il(x, y), x, tscrashlongitude),
      lat = if_else(in_il(x, y), y, tscrashlatitude),
      # crash date/time built from crashyr/crashmonth/crashday/crashhour
      # rather than the raw crashdate field, which is UTC-shifted and can
      # move late-evening crashes to the next calendar day.
      crash_date = make_date(crashyr, crashmonth, crashday),
      crash_dt   = crash_date + hours(crashhour),
      ped_bike = case_when(
        typeoffirstcrash == "Pedestrian"   ~ "Pedestrian",
        typeoffirstcrash == "Pedalcyclist" ~ "Bicycle",
        TRUE                               ~ NA_character_
      ),
      ksi = totalfatals > 0 | ainjuries > 0
    ) %>%
    select(-any_of("crash_hour"))

  crash <- crash_df %>%
    filter(in_il(lon, lat)) %>%
    st_as_sf(coords = c("lon", "lat"), crs = 4326, remove = FALSE)

  if (ped_bike_only) crash <- filter(crash, !is.na(ped_bike))

  st_transform(crash, crs_out)
}

idot_years <- 2022:2024
idot_dir   <- "data/boundaries"

crash <- process_crashes(idot_years, dir = idot_dir)
ped_bike <- process_crashes(idot_years, dir = idot_dir, ped_bike_only = TRUE)

st_write(crash, "output/idot_crashes_2022_2024.shp", delete_dsn = TRUE)
st_write(ped_bike, "output/idot_ped_bike_crashes_2022_2024.shp", delete_dsn = TRUE)
```

## Save

Save the assembled `sf` tables to a single `.RData` file. `phase01.qmd` loads this file to build an interactive data-exploration map; downstream analysis in `phase02.qmd` will read from these same tables (or their saved shapefiles under `output/`).

```{r}
#| label: save-dataprep

save(
  study_area, cc_boundary, cc_streets,
  all_intersections_sf, ped_fatalities,
  crash, ped_bike,
  file = "data/phase1_dataprep.RData"
)
```

## Outputs

- `data/boundaries/study_area.shp` — countywide study-area geography (Chicago community areas + suburban municipalities).
- `config/cc_run_list.csv` — per-study-area-unit run manifest (geoid, name, type, WKT polygon, wave).
- `data/intersections/raw/{geoid}.csv` — one file per study-area unit, raw consolidated intersection coordinates.
- `output/all_intersections.shp` — countywide consolidated-intersection layer.
- `output/all_streets.shp` — countywide street layer with `highway`, `maxspeed_raw`, `street_name`, `oneway`, and resolved `maxspeed_mph`/`maxspeed_source`.
- `output/ped_fatalities.shp` — Cook County pedestrian fatalities from FARS, 2010–2024.
- `output/idot_crashes_2022_2024.shp` / `output/idot_ped_bike_crashes_2022_2024.shp` — IDOT crash records, all crashes and the pedestrian/bicycle subset.
- `data/phase1_dataprep.RData` — `study_area`, `cc_boundary`, `cc_streets`, `all_intersections_sf`, `ped_fatalities`, `crash`, `ped_bike`, loaded by `phase01.qmd`'s interactive map.
