---
title: "Phase 1 — Dataset Development"
subtitle: "Street Design as a Public Health Determinant: Crosswalk Geometry and Pedestrian Injury Risk Across Cook County"
author:
  - name: "C. Scott Smith, PhD AICP"
    affiliations:
      - name: "DePaul University"
      - name: "Cook County Department of Public Health (CCDPH)"
    email: c.scott.smith@depaul.edu
  - name: "Zane Sufi"
    affiliations:
      - name: "Oak Park River Forest High School"
date: today
format:
  html:
    code-fold: true
    toc: true
    toc-depth: 2
    number-sections: false
    theme: cosmo
    embed-resources: true
  pdf:
    toc: true
    number-sections: true
editor: source
editor_options: 
  chunk_output_type: console
---

## Step 1 — Activate R Packages, Connect R to the OSMnx Python environment

```{r}
#| label: r-packages
#| eval: true
#| code-fold: true

library(sf)              # simple features: read, transform and map spatial data
library(reticulate)      # call Python (osmnx) from R
library(dplyr)           # data manipulation verbs (filter, mutate, select, joins)
library(purrr)           # functional iteration (map) over lists and files
library(readr)           # fast CSV reading/writing
library(glue)            # string interpolation
library(janitor)         # clean column names and tabulate (clean_names, tabyl)
library(cli)             # console alerts and progress messages
library(tigris)          # download Census TIGER/Line boundaries
library(units)           # unit-aware areas and lengths
library(leaflet)         # interactive web maps
library(leaflet.extras)  # addResetMapButton(), addFullscreenControl(), addGroupedLayersControl()
library(osmextract)      # read OpenStreetMap extracts
library(lubridate)       # date/time handling for crash records
library(stringr)         # string helpers (str_replace, str_detect)

# API key for leaflet maps
CARTO_API_KEY  <- Sys.getenv("CARTO_API_KEY",  unset = "cb1_31xh_1_06fb99adb466f0f5a0392d7d")


```

## Overview

This phase assembles every data layer needed to test the aims and hypotheses set out in [`phase00.html`](phase00.html): the countywide study-area geography, the street/intersection network, crash and fatality records, and socioeconomic/contextual covariates. Dataset development is in progress; the tables below record what is built, what is underway, and what has not yet started, organized by the research aim(s) each data layer supports.

Implementation (dataset-creation) code lives in [`dataprep.html`](dataprep.html), which saves the resulting `sf` tables to `data/phase01_dataprep.RData` — loaded below to build an interactive map for data exploration; this document describes what each step produces and why, not the implementation detail itself. `dataprep.qmd` supersedes the study-area/intersection/streets/FARS code originally developed in [`archive/Phase0.html`](archive/Phase0.html) (kept for reference) and the IDOT crash-processing logic prototyped in [`scripts/QMD/idot_crashes.html`](scripts/QMD/idot_crashes.html).

## Study Area Geography

Supports **Aims 1–4** (defines the unit of geography used to organize and aggregate every other data layer).

| Task | Description | Output | Status |
|---|---|---|---|
| Combine Chicago community areas and suburban municipalities | Build a single, internally consistent countywide geography spanning the City of Chicago's 77 community areas and suburban Cook County's incorporated municipalities (filtered to ≥4 sq mi to exclude small/fragmentary places) | Countywide study-area polygon layer (`study_area`), with a `type` field distinguishing community areas from municipalities | Complete |
| Reconcile boundary source and projection | Source boundaries from Cook County/Chicago health-atlas geographies; standardize to a common projected CRS for area and distance calculations | Projected study-area boundary file | Complete |

## Street Intersections

Supports **Aim 1** (defines candidate crossing locations for the segmentation pipeline) and **Aims 3–4** (defines the spatial units fatality/crash records are matched to).

| Task | Description | Output | Status |
|---|---|---|---|
| Clip OSM data to Cook County | Clip the Illinois OpenStreetMap extract from GeoFabrik to the Cook County boundary with `osmium extract` | Cook County OSM extract | Complete |
| Filter to the drivable street network | Keep drivable road types and remove parking aisles, driveways, private roads, airport features and indoor ways with `osmium tags-filter` | Cook County drive-network OSM file | Complete |
| Build street graph and consolidate intersections | Build one countywide street graph with OSMnx (`graph_from_xml()`), then merge OSM nodes within the default 10 m tolerance into single intersection points with `consolidate_intersections()`, excluding dead ends | Countywide intersection point layer (~170,000 points) | Complete |
| Assign intersections to study-area units | Spatially join intersection points to the study-area geography | `data/intersections/study_area_intersections.gpkg`, with `geoid`, `name` and `type` | Complete |

All processing uses a downloaded OSM file, so no online map service (Overpass API) is queried. See [`phase01_intersections.html`](phase01_intersections.html) for a full description of the methods.

## Crossing Distance (Segmentation Models)

Supports **Aim 1** directly — this is the study's core measured variable, and a prerequisite for Aims 2–4's crossing-distance exposure variable.

Per Aim 1a (`phase00.qmd`), imagery source and segmentation model selection is being validated on a sample of intersections before the countywide production run (Aim 1b). See [`phase01_crosswalk_models.html`](phase01_crosswalk_models.html) for the detailed comparison design and operationalization steps for all three candidate approaches (3-band Cook County orthoimagery, 4-band Cook County RGB+NIR orthoimagery, and Google Maps imagery).

| Task | Description | Output | Status |
|---|---|---|---|
| Imagery/model comparison (Aim 1a) | Measure crossing distance at a sample of intersections using all three imagery/model approaches; compare against manually measured ground truth | Comparison dataset and selected approach for full-scale production | Not started |
| Countywide production run (Aim 1b) | Apply the approach Aim 1a's results support, at full scale across all study-area intersections | Countywide crossing-distance dataset | Not started (gated on Aim 1a) |

## Streets and Speed Limits

Supports **Aims 3–4** (roadway functional class and speed limit are control variables in the fatality/injury models).

| Task | Description | Output | Status |
|---|---|---|---|
| Retrieve countywide street centerlines | Download Illinois streets via Geofabrik (`osmextract`) rather than per-municipality OSMnx calls, then clip to the study area | Countywide street centerline layer | Complete |
| Parse speed limits | Extract and standardize OSM `maxspeed` tags to miles per hour | Street segments attributed with posted speed limit | Complete |

## Pedestrian Fatalities (FARS)

Supports **Aim 3** directly.

| Task | Description | Output | Status |
|---|---|---|---|
| Download FARS records | Pull NHTSA FARS data for Illinois/Cook County, 2010–2024 | Raw FARS extract | Complete |
| Filter to pedestrian fatalities | Subset to pedestrian-involved fatalities with valid coordinates | Pedestrian-fatality point layer, with date/time-of-day attributes | Complete |

## Crash Records (IDOT)

Supports **Aim 4** directly (as the primary crash-injury data source; see *Open Items* below regarding EMS data).

| Task | Description | Output | Status |
|---|---|---|---|
| Stack yearly CSV exports | Combine 2022–2024 IDOT crash-report exports into a single table | Combined raw crash table | Complete |
| Clean coordinates and dates | Prefer in-state ArcGIS `x`/`y` coordinates, falling back to IDOT's reported longitude/latitude; reconstruct crash date/time from `crashyr`/`crashmonth`/`crashday`/`crashhour` rather than the raw `crashdate` field (which is UTC-shifted) | Cleaned crash table with valid point geometry | Complete |
| Identify pedestrian crashes and severity | Flag pedestrian-involved crashes from `typeoffirstcrash`; flag killed-or-seriously-injured (KSI) crashes from fatality/A-injury counts | `sf` point layer of pedestrian crashes, with severity and KSI flags | Complete |
| Extend to additional crash years | Add IDOT crash-year extracts beyond 2022–2024 as they become available, to increase statistical power for Aim 4 | Extended multi-year crash table | In progress |

## Socioeconomic and Contextual Data

Supports **Aim 2** directly (ADI, demographics) and provides control variables for **Aims 3–4**.

| Task | Description | Output | Status |
|---|---|---|---|
| Assemble Area Deprivation Index (ADI) | Download 2024 ACS variables via the `sociome`/`tidycensus` packages; aggregate 2020 census tracts (via a Chicago Data Portal tract-to-community-area lookup) to compute ADI by community area, and compute ADI directly by Census place for suburban municipalities | Community-area and suburban-municipality ADI tables | In progress |
| Assemble Census demographic covariates | Download ACS 5-year estimates (population density, housing tenure, vehicle ownership, age structure, poverty rate) | Municipality/community-area demographic covariate table | Not started |
| Assemble employment covariates | Download ACS/LEHD LODES data (employment density, commute mode share, jobs-housing balance) | Municipality/community-area employment covariate table | Not started |
| Assemble other contextual covariates | Incorporation year, land area, and other municipal/community-area attributes needed for Aim 2 comparability | Supplementary covariate table | Not started |
| Assemble transportation investment/obligations data | Extract CMAP's FFY2025 Annual Obligation Report (FHWA obligations, AC authorizations, FTA obligations) to Cook County municipality/project-type level, flagged for street-design relevance (signal, intersection, roadway, bike/ped, safety, bridge projects) | [`output/cook_county_transportation_obligations_ffy2025.xlsx`](output/cook_county_transportation_obligations_ffy2025.xlsx) — candidate exposure/control variable for Aims 2 and 4 (capital investment as a proxy for street-design change near crash locations) | Complete (FFY2025 only; extend with future years' reports) |

## Interactive Map (Data Exploration)

Loads a subset of `dataprep.qmd`'s saved tables (`data/phase01_dataprep.RData`) to explore the assembled data layers together: the Cook County boundary, the study-area geography, street intersections, Councils of Mayors boundaries, walkability (NTAscore), Area Deprivation Index (ADI) by study-area unit, CMAP 2020 land use, IDOT annual average daily traffic (AADT), and the two point layers used in Aims 3–4 (FARS pedestrian fatalities and IDOT pedestrian crashes). The street network isn't included here — it's dense enough to warrant its own view rather than competing with everything else on one map. This is an exploration view, not a finished figure — descriptive and statistical analysis of these layers is the subject of `phase02.qmd`.

```{r}
#| label: dataprep-map
#| eval: true
#| echo: true
#| code-fold: true

load("data/phase01_dataprep.RData")

CARTO_API_KEY <- Sys.getenv("CARTO_API_KEY", unset = "cb1_31xh_1_06fb99adb466f0f5a0392d7d")
ZOOM_THRESHOLD <- 14

bbox <- st_bbox(study_area)
type_pal <- colorFactor(c("#9e9e9e", "#616161"), domain = c("community area", "municipality"))

# ADI by study area unit: `adi` (data/census/adi_study_area.csv) is a plain
# table keyed by GEOID, not spatial -- join it onto study_area's polygons
# to map it. 
study_area_adi <- study_area %>%
  left_join(
    adi %>% mutate(GEOID = as.character(GEOID)) %>% select(GEOID, ADI),
    by = c("geoid" = "GEOID")
  )

walk_pal <- colorNumeric("Greens", domain = walkability$NTAscore, na.color = "transparent")
adi_pal  <- colorNumeric("YlOrRd", domain = study_area_adi$ADI, na.color = "transparent")

# Land use: conventional planning colors (APA LBCS-style -- yellow residential,
# red commercial, blue institutional, purple industrial, gray transportation)
lu_colors <- c(
  "Residential"                = "#FFFF00",
  "Commercial"                 = "#FF0000",
  "Institutional"              = "#0000FF",
  "Industrial"                 = "#A020F0",
  "Transportation & Utilities" = "#BEBEBE",
  "Agriculture"                = "#C8D88C",
  "Open Space"                 = "#4DAF4A",
  "Vacant"                     = "#E8DCC8",
  "Water"                      = "#7FB2E5",
  "Other / Not Classified"     = "#DDDDDD"
)

# AADT: line color and width both step up with traffic volume
aadt_map <- aadt %>% filter(!is.na(AADT), AADT > 0)
aadt_bins <- c(0, 5000, 15000, 30000, Inf)
aadt_pal <- colorBin(c("#fdd49e", "#fc8d59", "#d7301f", "#7f0000"),
                     domain = aadt_map$AADT, bins = aadt_bins)
aadt_weight <- c(1.5, 2.5, 3.5, 5)[cut(aadt_map$AADT, aadt_bins, labels = FALSE)]

exploration_map <- leaflet(options = leafletOptions(preferCanvas = TRUE)) %>%
  # boundary lines get their own pane above the default overlay pane (z 400)
  # so they stay on top of whichever thematic layer is switched on
  addMapPane("boundary_lines", zIndex = 450) %>%
  addTiles(
    urlTemplate = paste0(
      "https://basemaps.cartocdn.com/rastertiles/light_all/{z}/{x}/{y}.png?key=",
      CARTO_API_KEY
    )
  ) %>%
  addPolygons(
    data = study_area,
    fillColor = ~type_pal(type),
    weight = 1,
    opacity = 0.5,
    color = "#fff",
    fillOpacity = 0.5,
    label = ~ paste0(name, " (", type, ")"),
    group = "Study Area"
  ) %>%
  addPolygons(
    data = cog_boundaries,
    fill = FALSE,
    color = "#2b8cbe",
    weight = 2.5,
    opacity = 0.9,
    dashArray = "4",
    label = ~name,
    options = pathOptions(pane = "boundary_lines"),
    group = "Councils of Mayors"
  ) %>%
  addPolygons(
    data = walkability,
    fillColor = ~walk_pal(NTAscore),
    fillOpacity = 0.65,
    stroke = FALSE,
    label = ~ paste0("Walkability (NTAscore): ", round(NTAscore, 1)),
    group = "Walkability"
  ) %>%
  addPolygons(
    data = study_area_adi,
    fillColor = ~adi_pal(ADI),
    weight = 1,
    opacity = 0.5,
    color = "#fff",
    fillOpacity = 0.7,
    label = ~ paste0(name, " (", type, ") — ADI: ", round(ADI, 1)),
    group = "ADI"
  ) %>%
  # addPolygons(
  #   data = land_use,
  #   fillColor = ~unname(lu_colors[lu_category]),
  #   fillOpacity = 0.7,
  #   stroke = FALSE,
  #   label = ~lu_category,
  #   group = "Land Use (CMAP 2020)"
  # ) %>%
  addPolylines(
    data = aadt_map,
    color = ~aadt_pal(AADT),
    weight = aadt_weight,
    opacity = 0.9,
    label = ~ paste0(
      coalesce(na_if(ROAD_NAME, ""), na_if(MARKED_NAM, ""), "Unnamed segment"), " — AADT: ",
      format(AADT, big.mark = ","), " vehicles/day (", AADT_YR, ")"
    ),
    group = "AADT"
  ) %>%
  addPolygons(
    data = county_boundary,
    fill = FALSE,
    color = "#000",
    weight = 4,
    opacity = 1,
    options = pathOptions(pane = "boundary_lines"),
    group = "Cook County"
  ) %>%
  addCircleMarkers(
    data = intersections,
    radius = 2,
    stroke = FALSE,
    fillOpacity = 0.6,
    fillColor = "#333",
    group = "Intersections"
  ) %>%
  addCircleMarkers(
    data = ped_fatalities,
    radius = 4.5,
    stroke = TRUE,
    color = "#FFF",
    weight = 0.25,
    fillOpacity = 0.8,
    fillColor = "darkred",
    # label = ~ paste0("Pedestrian fatality (FARS) — ", fatality_date, ", ", fatality_time),
    group = "FARS Ped Fatalities"
  ) %>%
  addCircleMarkers(
    data = ped_crashes,
    radius = 3.5,
    stroke = FALSE,
    fillOpacity = 0.6,
    fillColor = "#3182bd",
    label = ~ paste0(ped, " crash — ", severity),
    group = "IDOT Ped Crashes"
  ) %>%
  # addLegend(
  #   "bottomleft",
  #   colors = c("#9e9e9e", "#2b8cbe", "#333", "darkred", "#3182bd"),
  #   labels = c("Study area", "Councils of Mayors", "Intersections", "FARS fatality", "IDOT pedestrian crash"),
  #   opacity = 1
  # ) %>%
  # addLegend(
  #   "topright",
  #   pal = walk_pal,
  #   values = walkability$NTAscore,
  #   title = "Walkability<br>(NTAscore)",
  #   opacity = 0.8
  # ) %>%
  # addLegend(
  #   "bottomright",
  #   pal = adi_pal,
  #   values = study_area_adi$ADI,
  #   title = "ADI<br>(relative)",
  #   opacity = 0.8
  # ) %>%
  addScaleBar(
    position = "bottomright",
    options = scaleBarOptions(metric = TRUE, imperial = TRUE)
  ) %>%
  fitBounds(bbox[["xmin"]], bbox[["ymin"]], bbox[["xmax"]], bbox[["ymax"]]) %>%
  addResetMapButton() %>%
  addFullscreenControl() %>%
  # Placeholder layer for the "None" radio option: an invisible point, so
  # selecting "None" switches every thematic layer off. Every name listed in
  # addGroupedLayersControl() must match a `group =` used above exactly, or
  # that entry silently does nothing.
  addCircleMarkers(
    lng = mean(c(bbox[["xmin"]], bbox[["xmax"]])),
    lat = mean(c(bbox[["ymin"]], bbox[["ymax"]])),
    radius = 0, stroke = FALSE, fillOpacity = 0,
    group = "None"
  ) %>%
  # Grouped layer control (leaflet.extras >= 2.1.0): boundaries and crashes
  # are checkboxes (any combination); the thematic layers are radio buttons
  # (one at a time) via exclusiveGroups, which must name the "Thematic" group.
  addGroupedLayersControl(
    overlayGroups = list(
      "Boundaries" = c("Cook County", "Study Area", "Councils of Mayors"),
      "Crashes" = c("FARS Ped Fatalities", "IDOT Ped Crashes"),
      "Thematic" = c("None", "ADI", "Walkability", "AADT")
    ),
    options = groupedLayersControlOptions(
      exclusiveGroups = "Thematic",
      groupCheckboxes = FALSE,
      groupsCollapsed = FALSE,
      collapsed = FALSE
    )
  ) %>%
  # Off by default (checked in the control to show): Councils of Mayors,
  # all thematic layers except "None", and both crash layers
  hideGroup(c("Councils of Mayors",
              "ADI", "Walkability", "AADT",
              "FARS Ped Fatalities", "IDOT Ped Crashes")) %>%
  htmlwidgets::onRender(
    sprintf(
      "function(el, x) {
          var map = this;
          var threshold = %d;
          var zoomGatedGroups = ['Intersections'];
          function toggleByZoom() {
            var show = map.getZoom() >= threshold;
            zoomGatedGroups.forEach(function(name) {
              var group = map.layerManager.getLayerGroup(name);
              if (!group) return;
              if (show) {
                if (!map.hasLayer(group)) map.addLayer(group);
              } else {
                if (map.hasLayer(group)) map.removeLayer(group);
              }
            });
          }
          map.on('zoomend', toggleByZoom);
          toggleByZoom();
        }",
      ZOOM_THRESHOLD
    )
  )

exploration_map
```

`Intersections` is deliberately left out of the layer control above (as in `Phase0.qmd`'s map) so it's purely zoom-gated rather than a manual checkbox, since it's the largest layer (~170,000 points countywide).

## Open Items

- **EMS injury data.** The original plan identified Cook County/regional EMS-reported pedestrian injury data as a possible input to Aim 4, pending a data-use agreement. IDOT crash records (above) are the working primary data source for Aim 4; EMS data will be incorporated only if access is secured, without blocking dataset development or downstream analysis.
- **Downstream phases.** Once the tables above are complete, dataset assembly and joins (study area + intersections + streets + crashes/fatalities + socioeconomic covariates) feed directly into `phase02.qmd` (data presentation and analysis).
