---
title: "Phase 0 Workflow: OSMnx Intersection Counts per Cook County Municipality"
subtitle: "An R/RStudio pipeline parallel to agupta01/crossing-distances (inference/osm_ingest.py)"
date: today
format:
  html:
    toc: true
    toc-depth: 2
    number-sections: true
    theme: cosmo
    embed-resources: true
editor: source
execute:
  eval: false
editor_options: 
  chunk_output_type: console
---

## Purpose

This document operationalizes one task from `plan.qmd`'s Phase 0 — *"Estimate imagery/tile volume needed for suburban Cook County using OSMnx intersection counts per municipality"* — as a concrete, runnable R/RStudio workflow.

It is written to be **parallel to**, not a reimplementation of, the intersection-extraction logic already used in the reference methodology repository ([`agupta01/crossing-distances`](https://github.com/agupta01/crossing-distances), specifically `inference/osm_ingest.py`). Rather than porting OSMnx's algorithms into native R, this workflow uses `reticulate` to call the real Python `osmnx` package from R, so the intersection-consolidation results are the *same* algorithm, not an approximation of it. R's role here is orchestration: loading the local Cook County shapefiles, looping over municipalities, and assembling the summary tables — the geometric work (`consolidate_intersections`, graph filtering) is delegated to `osmnx` itself.

Two things differ deliberately from the reference repo's approach, because this project is county-scale rather than a name-geocoded, single-city pull:

1. **Boundaries come from local shapefiles**, not `osmnx.graph_from_place()`'s Nominatim geocoding. Step 2 builds `cc_boundary.shp` (Cook County outline) and `cc_municipalities.shp` (municipal polygons) from Census TIGER/Line data via `tigris` and writes them under `data/boundaries/`, and the workflow calls `ox.graph_from_polygon()` per municipality against those instead.
2. **One run per municipality**, batched into "waves" — the same batching idea as the reference repo's `config/envs_100_cities.csv` (`Code`/`Wave` columns), just keyed to ~135 Cook County municipalities instead of ~100 US metro areas.

All code chunks below default to `eval: false` (see YAML header) so nothing runs unattended on open — review each chunk, then flip to `eval: true` (or knit interactively in RStudio) to run it.

------------------------------------------------------------------------

## How this parallels `osm_ingest.py`

| Step | This R workflow | Reference Python (`inference/osm_ingest.py`) |
|---|---|---|
| Study-area input | `sf::st_read()` on local shapefiles, per-municipality polygon | `place: str` argument, geocoded via Nominatim |
| Drive-network pull | `ox$graph_from_polygon(poly, network_type="drive", custom_filter=...)` via reticulate | `ox.graph_from_place(query=place, network_type="drive", custom_filter=...)` (lines 124–128) |
| Highway/service filter | Same `highway_subtypes` / `service_exclusion` lists, same custom_filter string | Lines 51–71, 127 |
| Projection | `ox$project_graph(G_drive)` | Line 130 |
| Intersection consolidation | `ox$consolidate_intersections(G_drive, rebuild_graph=FALSE, dead_ends=FALSE)` | Line 131–133 |
| Coordinate export | `.to_crs("EPSG:4326")$get_coordinates()` converted to an R tibble | Line 136–138 |
| Crosswalk-adjacency filter (optional refinement) | `ox$features_from_polygon()` + `gpd$sjoin_nearest()` within `RADIUS = 25` m | Lines 140–250 |
| Per-place batching | `config/cc_muni_run_list.csv` with a `wave` column | `config/envs_100_cities.csv` with `Code`/`Wave` columns |

Because Step "Intersection consolidation" calls the identical `osmnx` function, results should match what the Python pipeline would produce for the same polygon and `osmnx` version — pin and record that version (Step 1) so future re-runs stay comparable.

------------------------------------------------------------------------

## Prerequisites

**Inputs:** none pre-supplied — Step 2 builds `data/boundaries/cc_boundary.shp` and `data/boundaries/cc_municipalities.shp` itself from Census TIGER/Line data via `tigris`. If you already have official Cook County GIS boundary files you'd rather use instead, drop them at those same paths and skip the `tigris` download portion of Step 2.

**R packages:**

```{r}
#| label: r-packages
install.packages(c(
  "sf", "reticulate", "dplyr", "purrr", "readr", "glue",
  "janitor", "cli", "mapview"
))
```

**Python environment (via reticulate):** a conda or venv environment with `osmnx` installed. Match the version family used in the reference repo's Modal image (`gboeing/osmnx:latest`, see `inference/utils.py`) — after installing, print `ox$__version__` and record it alongside these results for reproducibility, rather than assuming a specific pinned version here.

**Before running the chunk below:** `reticulate::conda_create()`/`use_condaenv()` need a working **conda** binary (from Anaconda or Miniconda) — this is a separate installer from Python itself. Having Python already installed (e.g., Python 3.13 from python.org, the Microsoft Store, or elsewhere) does **not** satisfy this; `conda_create()` will always throw `Error: Unable to find conda binary. Is Anaconda installed?` regardless of what Python version is already on the machine, because it's specifically looking for `conda`, not `python`. Pick one of the two paths below:

-   **No conda installed at all (recommended default)** — let `reticulate` manage its own private Miniconda install, which won't touch or conflict with your existing Python 3.13:

    ```{r}
    #| label: install-miniconda-if-needed
    #| eval: false
    reticulate::install_miniconda()
    ```

    **On Windows in particular:** after this finishes, **restart your R session** (RStudio: Session → Restart R, or close and reopen RStudio) before running `conda_create()`. Reticulate caches its conda-path lookup for the life of the R session, so if you run `install_miniconda()` and then immediately retry `conda_create()` in the *same* session, you can still get the "unable to find conda binary" error even though the install succeeded. A fresh session picks up the new install.

-   **`install_miniconda()` reported success, but `use_condaenv()`/`conda_create()` still fail** — with a warning like `Supplied path is not a conda binary: '.../r-miniconda'` or `'.../r-miniconda' not found`. This means something is pointing `reticulate` at the miniconda **install directory** itself (the path `install_miniconda()` printed, e.g. `C:/Users/<you>/AppData/Local/r-miniconda`) instead of the `conda` **executable** inside it — those are not the same path, and `reticulate` needs the executable. The most common cause is having previously set `RETICULATE_CONDA` (per the next bullet) to that install-root path rather than a binary. Fix: clear the override and let `reticulate` auto-manage its own miniconda install, then restart R:

    ```{r}
    #| label: clear-bad-reticulate-conda-override
    #| eval: false
    Sys.getenv("RETICULATE_CONDA")  # check what's currently set, if anything
    Sys.unsetenv("RETICULATE_CONDA")
    # If Sys.getenv() above returned something even in a fresh session, also check
    # for a lingering RETICULATE_CONDA= line in usethis::edit_r_environ() and remove it.
    ```

    Then restart R and retry `reticulate::conda_binary()` (should now resolve to something like `.../r-miniconda/condabin/conda.bat` on Windows) before re-running `conda_create()`/`use_condaenv()`.

-   **Conda is installed, but `reticulate` can't find it** (common when conda isn't on the `PATH` RStudio sees). Point `reticulate` at it explicitly, using the actual path to your `conda` **executable** (not just the install folder), then restart R:

    ```{r}
    #| label: point-reticulate-at-conda
    #| eval: false
    # Find your conda binary from a terminal first: `which conda` (macOS/Linux) or `where conda` (Windows).
    # Common locations:
    #   macOS/Linux: ~/anaconda3/bin/conda, ~/miniconda3/bin/conda, /opt/anaconda3/bin/conda
    #   Windows:     C:/Users/<you>/Anaconda3/Scripts/conda.exe or C:/Users/<you>/Miniconda3/Scripts/conda.exe
    Sys.setenv(RETICULATE_CONDA = "/full/path/to/conda")
    ```

-   **Would rather not install conda at all** — use your existing Python 3.13 via a `venv`-based environment instead of a conda one (no conda required). This replaces the `conda_create()`/`conda_install()`/`use_condaenv()` calls throughout this document with their virtualenv equivalents:

    ```{r}
    #| label: python-env-setup-venv-alternative
    #| eval: false
    reticulate::virtualenv_create("osmnx-env", python = "3.13")
    reticulate::virtualenv_install(
      "osmnx-env",
      packages = c("osmnx", "geopandas", "shapely")
    )
    # then in Step 1, use:
    # reticulate::use_virtualenv("osmnx-env", required = TRUE)
    ```

    Caveat: Python 3.13 is very new, and `geopandas`'s underlying GDAL/GEOS/PROJ dependencies (via `fiona`/`pyproj`) sometimes lag behind the newest Python release in having prebuilt pip wheels on Windows. If `virtualenv_install()` fails trying to build one of these from source, that's what's happening — fall back to the conda path above (conda-forge reliably has these prebuilt) or point `virtualenv_create()` at an older installed Python (3.11/3.12) if you have one, rather than 3.13.

```{r}
#| label: python-env-setup
reticulate::conda_create("osmnx-env", packages = "python=3.11")
reticulate::conda_install(
  "osmnx-env",
  packages = c("osmnx", "geopandas", "shapely"),
  pip = TRUE
)
```

------------------------------------------------------------------------

## Step 1 — Connect R to the OSMnx Python environment

If `conda_create()` in Prerequisites already succeeded, this chunk should just work. If you're hitting a conda-related error here, see the troubleshooting bullets in the Prerequisites section above (a step-specific duplicate used to live here — consolidated there so there's one place to look, and so chunk labels aren't duplicated across the document).

```{r}
#| label: setup
library(sf)
library(reticulate)
library(dplyr)
library(purrr)
library(readr)
library(glue)
library(janitor)
library(cli)
library(tigris)
library(units)

reticulate::use_condaenv("osmnx-env", required = TRUE)

ox <- reticulate::import("osmnx")
gpd <- reticulate::import("geopandas")
shapely_wkt <- reticulate::import("shapely.wkt")

ox$settings$use_cache <- TRUE
ox$settings$log_console <- TRUE

cli::cli_alert_info("osmnx version: {ox[['__version__']]}")
```

------------------------------------------------------------------------

## Step 2 — Create Cook County Geometry Layers

```{r}
#| label: load-boundaries
# download latest counties data from US Census TIGER/Line data
counties_il <- counties(state="17", cb=TRUE)

# reformat data, standardize column names
cc_boundary <- counties_il %>%
  filter(GEOID == "17031") %>% 
  st_transform(3435) %>%
  st_as_sf() %>%
  select(geoid_county = GEOID,
         name_county = NAME,
         area_land = ALAND,
         area_water = AWATER) %>%
  mutate(area_land = set_units(area_land,m^2),
         area_water = set_units(area_water,m^2),
         area_land_sqmi = set_units(area_land, mi^2),
         area_water_sqmi = set_units(area_water,mi^2),
         area_sqmi = area_land_sqmi + area_water_sqmi) %>%
  select(-c(area_land,area_water))

# write cook county boundary to data/boundaries folder
st_write(obj = cc_boundary, dsn = "data/boundaries/cc_boundary.shp")

# download latest places data from US Census TIGER/Line data
il_places <- places(state="17", cb=TRUE) %>% 
  st_transform(3435) %>%
  st_as_sf() 

# cook county geometry
cc_geom <- st_geometry(cc_boundary)

# Fully within the boundary
places_in_cc <- il_places %>%
  filter(lengths(st_within(geometry, cc_geom)) > 0)

# Any overlap
places_touching_cc <- il_places %>%
  filter(lengths(st_intersects(geometry, cc_geom)) > 0)

# Centroid inside the boundary
places_by_centroid <- il_places %>%
  filter(lengths(st_within(st_centroid(geometry), cc_geom)) > 0)

# rename and format places
cc_municipalities <- places_by_centroid %>% 
  select(geoid_place = GEOID,
         name_place = NAME,
         area_land = ALAND,
         area_water = AWATER) %>%
  mutate(area_land = set_units(area_land,m^2),
         area_water = set_units(area_water,m^2),
         area_land_sqmi = set_units(area_land, mi^2),
         area_water_sqmi = set_units(area_water,mi^2),
         area_sqmi = area_land_sqmi + area_water_sqmi) %>%
  select(-c(area_land,area_water))

# write cook county boundary to data/boundaries folder
st_write(obj = cc_municipalities, dsn = "data/boundaries/cc_municipalities.shp")

# read cook county boundary
cc_boundary <- st_read("data/boundaries/cc_boundary.shp") %>%
  st_make_valid() %>%
  st_transform(4326)  # osmnx expects unprojected lat/lon polygons

# read municipal layer
cc_municipalities <- st_read("data/boundaries/cc_municipalities.shp") %>%
  st_make_valid() %>%
  st_transform(4326)

# Confirm the attribute field(s) that identify each municipality before proceeding
names(cc_municipalities)
```

The `select()` step above names these fields `name_place`/`geoid_place`, but the Shapefile/DBF format truncates field names to 10 characters, so after the `st_write()`/`st_read()` round-trip above they come back truncated — confirmed here as `nam_plc`/`ged_plc` (re-check with `names(cc_municipalities)` above if you regenerate the shapefiles, since GDAL's truncation/deduplication isn't always the same abbreviation):

```{r}
#| label: normalize-muni-fields
MUNI_NAME_FIELD <- "nam_plc"
MUNI_ID_FIELD   <- "ged_plc"

cc_municipalities <- cc_municipalities %>%
  rename(muni_name = all_of(MUNI_NAME_FIELD),
         muni_id = all_of(MUNI_ID_FIELD)
         )
```

**Coverage decision (see `plan.qmd`, Open Decisions):** if unincorporated Cook County is in scope for this run, derive it as the remainder of the county boundary after removing all municipalities, and give it its own pseudo-municipality row so it flows through the same pipeline:

```{r}
#| label: unincorporated-remainder
unincorporated <- st_difference(cc_boundary, st_union(cc_municipalities)) %>%
  st_collection_extract("POLYGON") %>%
  mutate(muni_id = "UNINC", muni_name = "Unincorporated Cook County") %>%
  select(muni_id, muni_name, geometry)

study_area <- bind_rows(
  cc_municipalities %>% select(muni_id, muni_name, geometry),
  unincorporated  # omit this line if unincorporated areas are out of scope
)
```

------------------------------------------------------------------------

## Step 3 — Build the per-municipality run list (parallel to `envs_100_cities.csv`)

```{r}
#| label: build-run-list
N_PER_WAVE <- 12  # tune based on how much Overpass load you want per batch

muni_run_list <- study_area %>%
  mutate(wkt = st_as_text(geometry)) %>%
  st_drop_geometry() %>%
  arrange(muni_name) %>%
  mutate(wave = ceiling(row_number() / N_PER_WAVE))

dir.create("config", showWarnings = FALSE)
write_csv(muni_run_list, "config/cc_muni_run_list.csv")
```

Each row is one municipality's polygon (as WKT) plus a `wave` number — directly analogous to a row in the reference repo's `config/envs_100_cities.csv`, just keyed to Cook County municipalities instead of metro-area airport codes.

------------------------------------------------------------------------

## Step 4 — Define the per-municipality intersection extraction function

Same `highway_subtypes` / `service_exclusion` filter and the same `custom_filter` string construction as `osm_ingest.py` lines 51–71 and 127:

```{r}
#| label: filter-constants
highway_subtypes <- c(
  "primary", "primary_link", "unclassified", "residential",
  "secondary", "secondary_link", "service", "tertiary",
  "tertiary_link", "trunk", "trunk_link", "busway",
  "construction", "corridor"
)
service_exclusion <- c("parking_aisle", "driveway")

drive_filter <- glue(
  '["highway"~"{paste(highway_subtypes, collapse = "|")}"]',
  '["service"!~"{paste(service_exclusion, collapse = "|")}"]',
  '["access"!~"private"]["aeroway"!~".*"]["indoor"!~"yes"]'
)
```

```{r}
#| label: intersection-function
#' Extract OSMnx-consolidated drive-network intersections for one municipality
#'
#' @param muni_id,muni_name identifiers carried through to the output rows
#' @param wkt WKT polygon string for the municipality (EPSG:4326)
#' @param retain_all if TRUE, keep disconnected network components (osmnx
#'   default is FALSE, matching osm_ingest.py; consider TRUE for small/exurban
#'   municipalities where a fragmented street network could otherwise be
#'   under-counted)
#' @param tolerance override osmnx's default consolidate_intersections
#'   tolerance (left NULL here to match osm_ingest.py, which does not
#'   override it)
get_muni_intersections <- function(muni_id, muni_name, wkt,
                                    retain_all = FALSE, tolerance = NULL) {
  poly <- shapely_wkt$loads(wkt)

  G_drive <- tryCatch(
    ox$graph_from_polygon(
      poly,
      network_type = "drive",
      custom_filter = drive_filter,
      retain_all = retain_all
    ),
    error = function(e) {
      cli::cli_alert_warning("{muni_name}: {conditionMessage(e)}")
      NULL
    }
  )
  if (is.null(G_drive)) return(NULL)

  G_drive <- ox$project_graph(G_drive)

  consolidate_args <- list(G = G_drive, rebuild_graph = FALSE, dead_ends = FALSE)
  if (!is.null(tolerance)) consolidate_args$tolerance <- tolerance
  intersections <- do.call(ox$consolidate_intersections, consolidate_args)

  intersections_gdf <- gpd$GeoDataFrame(
    geometry = intersections, crs = intersections$crs
  )$to_crs("EPSG:4326")

  coords <- reticulate::py_to_r(intersections_gdf$get_coordinates())
  coords %>%
    as_tibble() %>%
    rename(lon = x, lat = y) %>%
    mutate(muni_id = muni_id, muni_name = muni_name)
}
```

------------------------------------------------------------------------

## Step 5 — (Optional refinement) Filter to intersections near a pedestrian crossing

Raw drive-network intersection counts overstate the true tile requirement — many suburban intersections have no marked or unmarked pedestrian crossing at all. `osm_ingest.py` (lines 140–250) filters intersections down to those within `RADIUS = 25` m of an OSM-tagged crossing before treating them as tile targets. The same join, translated:

```{r}
#| label: crossing-adjacency-filter
RADIUS_M <- 25.0  # matches inference/utils.py RADIUS

get_muni_crossing_adjacent_intersections <- function(muni_id, muni_name, wkt) {
  poly <- shapely_wkt$loads(wkt)

  crosswalk_features <- tryCatch(
    ox$features_from_polygon(
      poly, tags = list(highway = "crossing", crossing = TRUE, disused = FALSE)
    ),
    error = function(e) NULL
  )
  if (is.null(crosswalk_features) || nrow(reticulate::py_to_r(crosswalk_features)) == 0) {
    return(NULL)  # no tagged crossings in this municipality
  }

  intersections_result <- get_muni_intersections(muni_id, muni_name, wkt)
  if (is.null(intersections_result)) return(NULL)

  intersections_sf <- st_as_sf(intersections_result, coords = c("lon", "lat"), crs = 4326) %>%
    st_transform(3857)

  crossings_sf <- reticulate::py_to_r(crosswalk_features$to_crs("EPSG:3857")) %>%
    st_as_sf()

  nearest <- st_is_within_distance(intersections_sf, crossings_sf, dist = RADIUS_M)
  intersections_result[lengths(nearest) > 0, ]
}
```

Run this as a second pass over Step 4's output for municipalities where a more precise tile-volume estimate is worth the extra Overpass query; the raw count from Step 4 is a reasonable, faster upper-bound estimate for initial Phase 0 sizing.

------------------------------------------------------------------------

## Step 6 — Run across all municipalities, wave by wave

Unlike the GPU segmentation stage (Modal, embarrassingly parallel), this step queries the Overpass API underneath `osmnx`. The shared public Overpass endpoint has usage limits and etiquette expectations — high-concurrency requests risk 429s/timeouts and are discouraged. Process **sequentially within each wave**, with a short pause between municipalities, and checkpoint so a failed run can resume without redoing completed municipalities. If higher throughput is genuinely needed, stand up a private/self-hosted Overpass instance rather than hammering the public one.

```{r}
#| label: run-waves
out_dir <- "data/intersections/raw"
dir.create(out_dir, recursive = TRUE, showWarnings = FALSE)

run_wave <- function(wave_df, pause_sec = 1) {
  purrr::pwalk(wave_df, function(muni_id, muni_name, wkt, wave, ...) {
    out_file <- file.path(out_dir, glue("{muni_id}.csv"))
    if (file.exists(out_file)) return(invisible(NULL))  # resume support

    cli::cli_progress_step("Processing {muni_name}")
    result <- get_muni_intersections(muni_id, muni_name, wkt)
    if (!is.null(result)) readr::write_csv(result, out_file)
    Sys.sleep(pause_sec)
  })
}

muni_run_list %>%
  group_split(wave) %>%
  walk(run_wave)
```

------------------------------------------------------------------------

## Step 7 — Combine and summarize counts

```{r}
#| label: summarize
intersection_files <- list.files(out_dir, full.names = TRUE)
all_intersections <- map_dfr(intersection_files, read_csv, show_col_types = FALSE)

muni_summary <- all_intersections %>%
  count(muni_id, muni_name, name = "intersection_count") %>%
  left_join(st_drop_geometry(study_area), by = c("muni_id", "muni_name"))

county_total <- sum(muni_summary$intersection_count)
chicago_benchmark <- 80000
scale_factor <- county_total / chicago_benchmark

cli::cli_alert_info(
  "Suburban Cook County: {county_total} intersections (~{round(scale_factor, 2)}x the Chicago benchmark of {chicago_benchmark} tiles)"
)

dir.create("output", showWarnings = FALSE)
write_csv(muni_summary, "output/cc_intersection_counts_by_municipality.csv")
```

This total feeds directly into `plan.qmd`'s Phase 0 deliverable ("cloud vendor selection and cost-per-tile benchmark") and the Cloud Compute Strategy scenarios already documented there (Modal GPU-tier/concurrency choices sized off this tile-count estimate).

------------------------------------------------------------------------

## Step 8 — QA / sanity checks

-   **Visual check:** map intersection points over municipality polygons to confirm no municipality came back empty and points fall inside their boundary.

    ```{r}
    #| label: qa-map
    library(mapview)
    all_points_sf <- st_as_sf(all_intersections, coords = c("lon", "lat"), crs = 4326)
    mapview(study_area, alpha.regions = 0.1) + mapview(all_points_sf, cex = 1)
    ```

-   **Outlier review:** flag municipalities whose intersection count looks implausible relative to land area or population (from the municipality attribute table) for manual review — this feeds forward into Phase 1's "Document OSM crossing-coverage completeness by municipality" validation task in `plan.qmd`.

-   **Boundary double-counting:** `ox.graph_from_polygon()` clips the network to each polygon, but nodes sitting exactly on a shared border between two municipalities could in principle appear in both extractions. If a single county-wide (not just per-municipality) point dataset is assembled later in Phase 1, snap coordinates to a small grid (e.g., 1 m) and de-duplicate before combining.

-   **Disconnected components:** the default `retain_all = FALSE` (matching `osm_ingest.py`) keeps only the largest connected component per municipality. For small or exurban municipalities, a fragmented street network could lose a disconnected pocket of intersections — spot-check a few such municipalities with `retain_all = TRUE` and compare counts before deciding which setting to use county-wide.

------------------------------------------------------------------------

## Outputs

-   `config/cc_muni_run_list.csv` — per-municipality run manifest (id, name, WKT polygon, wave).
-   `data/intersections/raw/{muni_id}.csv` — one file per municipality, raw consolidated intersection coordinates.
-   `output/cc_intersection_counts_by_municipality.csv` — municipality-level summary table (intersection count, joined attributes) and the county-wide total used for tile-volume and cloud-cost sizing in Phase 0.

------------------------------------------------------------------------

## Note on fidelity and alternatives

Because Step 4/5 call the actual `osmnx` Python functions through `reticulate`, results should match what the Python pipeline would produce for the same polygon and `osmnx` version — R here is orchestration, not a reimplementation of OSMnx's algorithms.

A pure-R alternative (no Python/reticulate) is possible using `osmdata` (an Overpass API wrapper) plus `sf`/`igraph`/`sfnetworks` for network handling, but OSMnx's `consolidate_intersections` merging algorithm would need to be reimplemented from scratch (e.g., via buffering, unioning, and centroid clustering of nearby nodes), and there's no guarantee it reproduces identical counts. Not recommended if numerical parity with the reference methodology matters, which it does here since Phase 0's tile-volume estimate is meant to be directly comparable to the Chicago benchmark produced by the Python pipeline.
